---
format: html
bibliography: bibliography/Boot-Err_Tutorial.bib
csl: bibliography/apa-6th-edition.csl
---

Lukas J. Beinhauer *University of Erfurt* ORCID: https://orcid.org/0000-0001-9841-3089

Jens H. Fünderich *Ludwig Maximilian University of Munich* ORCID: https://orcid.org/0000-0002-7185-9248

Frank Renkewitz *University of Erfurt* ORCID: https://orcid.org/0000-0001-8072-6802

All communications should be directed to the corresponding author:\
Lukas J. Beinhauer\
Nordhäuser Str. 63, 99089 Erfurt, DE\
lukas.beinhauer\@uni-erfurt.de


#### Abstract

Researchers frequently administer the same psychological scale across different contexts for various reasons, such as online/offline, different cultural backgrounds or language translation. In such situations, assessing whether measurement quality is consistent across administrations is an important step. Several established methods address this question, making use of different interpretations of measurement quality. Reliability Generalization Meta-Analysis, residual invariance testing, and Boot-Err each answer a subtly different question about measurement quality, and applying the wrong tool to a given research problem can lead to misleading conclusions. This text has two goals. First, we clarify what each method actually does, identify the research contexts each is best suited for, and detailling how the three approaches are complements rather than competitors. Reliability Generalization Meta-Analysis addresses a ratio-based question about the signal-to-noise ratio of a scale across groups. Residual invariance testing asks whether a measurement model holds equivalently across a small, well-defined set of groups. Boot-Err quantifies how much random measurement error varies across a potentially large number of independent administrations and whether that variation is large enough to matter. Second, because accessible tutorials exist for the first two methods but not for Boot-Err, we provide a step-by-step implementation guide in R, illustrated using the HEXACO-Conscientiousness scale administered across 19 independent laboratories. Applying Boot-Err to this dataset, we find statistically significant heterogeneity in error score variance across laboratories. However, the overwhelming majority of heterogeneity in total score variance can not be attributed to these differences in random measurement error, but to differences in true score variance instead.

# Measuring measurement quality: A critical comparison of three approaches to random error

Researchers frequently administer the same psychological scale across different contexts or populations. A scale may be translated into another language, administered online instead of in person, or completed by participants with different educational or cultural backgrounds (for selected examples see Klein et al. [-@ManyLabs1], Klein et al. [-@ManyLabs2], Verschuere et al. [-@RRRMazar]). In such situations, an important question arises: Is measurement quality consistent across these administrations?

At first glance, this question appears straightforward and several established methods seem to address it. Reliability Generalization Meta-Analysis [@vacha-haase1998], residual invariance testing [@meredith1993; @vandenberg2000] and Boot-Err [@beinhauer2025] all speak to the question of whether measurement quality differs across contexts. Because they share this surface-level concern, they might seem interchangeable. However, each method answers a subtly but importantly different question. Applying the wrong tool to a given research problem can lead to misleading conclusions.

The goal of this text is threefold: (1) to clarify what each method actually does, (2) to identify the research contexts each method is best suited for, and, because accessible tutorials already exist for Reliability Generalization Meta-Analysis and residual invariance testing, (3) provide a hands-on tutorial for Boot-Err, where such guidance is currently lacking. The text itself is structured along two major sections. In the first section "Background", arguments regarding the goals (1) and (2) are discussed. The second section "Tutorial", employing a data example and R-code, is structured to meet goal (3).

This text is written for researchers who are already familiar with the fundamental concepts of classical test theory and confirmatory factor analysis, and who have working knowledge of R. No prior experience with meta-analysis or bootstrapping is required, as these are introduced as needed throughout. Readers who are new to reliability coefficients or measurement invariance testing may find it helpful to first consult McNeish [-@mcneish2018] or Putnick \& Bornstein [-@putnick2016] before proceeding.

## Background

Throughout this text, we focus on random measurement error as a central component of measurement quality. Random measurement error refers to unsystematic fluctuations in observed scores around the values they would take on in the absence of measurement imprecision [@lord1974]. Unlike systematic error, which shifts observed scores in a particular direction, random error introduces additional, unsystematic variation sometimes referred to as noise. On average, observed scores may still be accurate, but individual measurements vary unpredictably around their underlying true values.

If the same instrument is administered in two different contexts, differences in random measurement error imply differences in measurement quality [@meredith1993; @lord1974]. However, the three methods discussed here do not address this in the same way. Instead, they reflect different perspectives on what it means for measurement quality to be equivalent, and they are therefore suited to different research questions.

### Reliability generalization meta-analysis

The first, most intuitive and widely used approach is to compare reliability coefficients across administrations of the same scale [e.g. @vacha-haase1998; @shaw2020]). The conceptual idea behind reliability coefficients can be described in terms of classical test theory [@lord1974]. Classical test theory implies that each score $X$ is made up of true score $T$ and error score $E$:

$$ X = T + E$$ {#eq-CTT}

Reliability coefficients, such as Cronbach's alpha [@cronbach1951], McDonald's Omega [@mcdonald1999] or Guttmann's Lambda IV [@guttman1945] quantify the proportion of true score variance present in the scores:

$$\rho_{XX'} = \frac{\sigma^2_T}{\sigma^2_X}$$ {#eq-reliability}

where $\rho_{XX'}$ is score reliability, $\sigma^2_T$ is the true score variance and $\sigma^2_X$ is the total score variance, made up of true and error score variance ($\sigma^2_X = \sigma^2_T + \sigma^2_E$, where $\sigma^2_E$ is the error score variance).

This true score variance $\sigma^2_T$ describes the variation in the scores, that can be attributed to genuine differences in the underlying factor being measured. Error score variance $\sigma^2_E$ describes the variation in the scores, that can be attributed to random, unsystematic noise. It quantifies the random measurement error that is introduced by the instrument's inability to perfectly measure the underlying factor. Reliability coefficients thereby express the proportion of observed score variance that cannot be attributed to random measurement error, and takes on values between 0 and 1 [@lord1974].

Reliability Generalization Meta-Analysis extends this idea by formally testing whether observed differences in score reliability exceed what would be expected from sampling error alone [@vacha-haase1998]. Because coefficients derived from samples are always prone to sampling variability, one can not simply compare point estimates across administrations [@feldt1987; @cohen1994; @beretvas2003]. Meta-analytic techniques account for this sampling error, derive an estimate of heterogeneity and test whether it is larger than 0 [@borenstein2009; @hunter2015]. Meta-analytic heterogeneity refers to variance that exceeds the variation that sampling fluctuation alone would produce. If significant heterogeneity is detected, this indicates that reliability differs across administrations in a way, that is likely not attributable to chance alone [@sánchez-meca2012; @rodriguez2006; @beretvas2003].

**What question does it answer?** Reliability Generalization Meta-Analysis addresses a ratio-based question: in which contexts is the scale relatively more informative, in the sense that a greater proportion of observed score variance reflects true differences among individuals?

**When is this the right tool?** This approach is well suited when the research question concerns the *signal-to-noise ratio* of a scale across populations that are expected to differ in the spread of the underlying construct. Consider an ability test developed and validated in a group of university students. This describes a common practice driven by convenience, that persists despite growing recognition that student samples are not representative of the general population [@petrutiu2026]. University students are a relatively homogeneous group with respect to ability. Students likely show less variation in ability than the general population. 

In this context, the relative amount of measurement error, the signal-to-noise ratio as assessed through score reliability is of central interest in terms of measurement quality. If score reliability for the general population group is approximately equal to score reliability for the university student group, then the ability test discriminates equally well between the levels of ability in both groups. If it differs, the reasons for those differences and their practical implications warrant further investigation [e.g. @gnambs2015]. Reliability Generalization Meta-Analysis can reveal this directly, informing decisions about whether the test usefully differentiates among individuals within both contexts.

**Limitations** Because score reliability is a ratio, its value is affected by differences in true score variance just as much as by differences in error score variance.

For simplicity, the examples that follow compare only two groups. In practice, however, researchers are often interested in comparing score reliability across many administrations, populations, or study contexts. When only two groups are involved, differences in reliability coefficients can be evaluated directly using procedures such as Feldt's test [@feldt1987]. The two-group examples presented here are intended solely to illustrate the underlying logic of how reliability depends on both true-score variance and error variance. The same principles generalize directly to settings involving multiple groups.

Consider the following hypothetical for our ability test example. Suppose the true reliability of the test is .67 in both groups:

$$\rho_{xx'\ students} = \frac{2}{2 + 1} = .67 \qquad \rho_{xx'\ general} = \frac{3}{3 + 1.5} = .67$$ {#eq-RGMA2}

In this pattern, identical score reliability coexists with differences in true and error score variance. True and error score variance retain the same proportion in both groups. This means that the relative amount of measurement error in the observed scores is identical across both groups. This is the core mechanism, why the test can be used to differentiate equally well between different ability levels within each of the groups.

If, on the other hand, the goal is to compare ability levels *between* the two groups, the context changes. For instance, we might ask whether university students score higher than the general population. Now what matters is not the signal-to-noise ratio within each group, but whether individual measurements carry the same degree of precision across groups. The difference in error score variance ($1$ in the student group, $1.5$ in the general population) means that measurements in the general population group are less precise: observed scores deviate more strongly from individuals' true ability levels than they do in the student group. Reliability coefficients, being ratios, cannot detect this asymmetry. In both groups the true reliability coefficient is $\rho_{xx'} = .67$. This is the limitation that motivates moving from a relative to an absolute perspective on measurement. More broadly, when the research question concerns the precision of individual measurements, rather than how well a scale discriminates within a given group, the absolute magnitude of error score variance is the more informative quantity.

Consequently, Reliability Generalization Meta-Analysis answers a ratio-based question about measurement quality, not a question about differences in the absolute magnitude of measurement error.

### Measurement invariance testing

A second perspective is offered by measurement invariance testing, grounded in confirmatory factor analysis [@meredith1993; @vandenberg2000] and typically implemented using structural equation modelling software such as lavaan [@rosseel2012] or Mplus [@muthen2017]. Rather than summarising measurement quality as a single coefficient per group, this framework models the relationship between latent factors and observed item responses and tests, whether this relationship holds equivalently across groups.

Measurement invariance testing proceeds through a series of increasingly constrained models, typically four levels [@vandenberg2000; @vandeschoot2012; @meredith1993]:

-   Configural invariance: The same factor structure holds across all groups.

-   Metric invariance: Factor loadings are identical across groups.

-   Scalar invariance: Intercepts are identical, meaning systematic differences in observed scores across groups can be attributed to differences in latent factors.

-   Residual invariance: Residual variances describing variation in item scores not explained by the latent factor are identical across groups.

The last level is of greatest relevance here. In classical test theory terms, residual variance can be interpreted as item-level error score variance (with some caveats, see Raykov [-@raykov1997] or Kline [-@kline2023]). The test for residual invariance evaluates whether these variances are equal across groups by comparing model fit between an unconstrained model (residual variances freely estimated) and a constrained model (residual variance held equal) [@meredith1993; @vandenberg2000; @vandeschoot2012]. A statistically significant $\Delta\chi^2\ (p < .05)$ suggests that the constraint meaningfully worsens model fit, indicating a lack of residual invariance. Because chi-square tests are sensitive to sample size, researchers typically supplement this with changes in approximate fit indices. Conventional guidelines suggest that small changes (e.g. $\Delta CFI \leq .1$, $\Delta RMSEA \leq .015$, $\Delta SRMR \leq .01$) indicate that invariance holds. However, these thresholds are heuristic rather than strict rules and vary across sources and simulation conditions [e.g. @steenkamp1998; @vandenberg2000; @meade2008; @cheung2002].

**What question does it answer?** Residual invariance testing addresses a *confirmatory, model-based* question: does the measurement model hold equivalently across groups, in the sense that item-level error variances are statistically indistinguishable?

**When is this the right tool?** This approach is well suited when the primary goal is to verify that the statistical requirements for group comparison are met. A common example is cross-cultural research on a personality dimension and needing to establish that scale scores are genuinely comparable. It is particularly appropriate when working with a smaller, well-defined set of groups and when the question is fundamentally confirmatory, whether the model holds. We return to the example above (@eq-RGMA2), where score reliability coefficients were identical despite differing error score variances. For that example, we established that differences in ability may be meaningfully compared within groups, as we learned from Reliability Generalization Meta-Analysis. However, in order to ensure that ability levels may be compared across groups, we need to assess whether the absolute magnitude in error score variance is equivalent across groups. Residual invariance testing, given adequate power, would correctly flag the difference in error score variance that reliability coefficients would have missed.

**Limitations** Importantly, the test for residual invariance generates sensible results only if the previous levels of invariance are established [@meredith1993; @steenkamp1998; @vandeschoot2012]. If an earlier level could not be confirmed, for example, if scalar invariance does not hold, results from the test for residual invariance cannot be interpreted as evidence about differences in random measurement error, since the necessary preconditions for that interpretation are not met.

If assumptions are met, while residual invariance testing can detect whether error score variances differ, it does not quantify the magnitude of those differences. Whether the discrepancy in error score variance is .5 or 17, the test result is qualitatively the same, as invariance either does or does not hold. When the scale is administered across a large number of contexts, and the research question concerns not merely whether measurement quality varies, but how *much*, residual invariance testing is not sufficient to derive a meaningful answer.

### Boot-Err - inspecting differences in random measurement error

Boot-Err addresses this gap. Like residual invariance testing, it conceptualises measurement quality in terms of the absolute magnitude of measurement error rather than its total variance [@beinhauer2025]. Unlike residual invariance testing, it provides an estimate of the *magnitude of heterogeneity* in error score variance across administrations, along with a formal statistical test.

The key identity underlying Boot-Err is that error score variance can be derived from observed score variance and a score reliability estimate:

$$\hat{\sigma}^2_E = (1 - \hat{\rho}_{xx'})\ \hat{\sigma}^2_X$$ {#eq-errorvar}

Bootstrapping is used to approximate the sampling variability of this estimate, expressed as a standard error. After applying appropriate variance-stabilising transformations [@bonett2002], error score variance estimates from multiple administrations are entered into a random-effects meta-analysis (with subsequent back-transformation [@beinhauer2025]). The random-effects framework is essential here: unlike fixed-effects meta-analysis, which assumes zero heterogeneity in the population, random-effects models explicitly accommodate and estimate the between-administration variance that Boot-Err is designed to quantify [@borenstein2009; @hunter2015]. The result is both a test of whether differences in error score variance are statistically significant and an estimate of their magnitude, expressed as heterogeneity $\tau^2_{\sigma^2_E}$.

**What question does it answer?** Boot-Err addresses a *quantitative* question about measurement error: how much does random measurement error vary across administrations, and is that variation large enough to matter?

**When is this the right tool?** Boot-Err is best suited when the same scale has been administered across a large number of contexts, such as across laboratories in a multi-lab replication project, or across cultural groups in a large-scale cross-national study. The research question largely pertains to the *magnitude* of variability in measurement quality. This is relevant, for example, when trying to understand how much variation in measurement error contributes to heterogeneity in effect sizes across labs, or when deciding whether differences in measurement error are large enough to complicate pooling of data across sites. Because Boot-Err operates on error score variance directly, rather than reliability ratios, it is not confounded by differences in true score variance across groups. Differences in random measurement error can still be detected and quantified. Unlike residual invariance testing, Boot-Err is designed for settings with many independent administrations and yields a single heterogeneity estimate with a formal test, rather than group-wise comparisons of residual variances.

Returning to the example in @eq-RGMA2, where reliability coefficients were equal but error score variances differed ($\sigma^2_{E, students} = 1$, $\sigma^2_{E, general} = 1.5$) - Boot-Err, if appropriately powered, would detect this difference and estimate the heterogeneity as $\tau^2_{\sigma^2_E} = .065$ ($\tau_{\sigma^2_E} = .25$). Given substantive knowledge about the scale and the expected magnitude of true and error score variance, researchers can evaluate whether this degree of variation has meaningful practical implications for group comparisons or effect sizes. In the chapter *Bonus-Step* of the following tutorial, we highlight perspectives on how such a question can be discussed and framed.

**Limitations** Boot-Err relies on several assumptions that should be made explicit. First, error score variance is derived from reliability estimates, which themselves depend on specific measurement models and assumptions such as unidimensionality and appropriate factor structure [@mcdonald1999; @raykov1997; @sijtsma2009]. Consequently, Boot-Err does not estimate a model-free quantity of random error, but rather the error variance implied by a given reliability framework. Second, differences in estimated error variance may reflect not only changes in measurement precision but also differences in item covariance structure, dimensionality, or model misfit across contexts. Third, while Boot-Err quantifies heterogeneity in error variance, there are currently no established benchmarks for interpreting the practical magnitude of such heterogeneity, leaving substantive interpretation to domain-specific judgment. We do not attempt to provide practical benchmarks, as these are bound to be largely domain-specific. However, at the end of the following tutorial, we will discuss how researchers can explore the relevance of error score variance heterogeneity for their specific use case.

### Summary and scope

The three methods address three distinct questions, summarised in @tbl-summary. Reliability Generalization Meta-Analysis assesses whether the *proportion* of true score variance differs across contexts, a ratio-based question about relative signal quality. Residual invariance testing assesses whether the measurement model holds equivalently, a confirmatory question about whether item-level error variances are statistically indistinguishable across a defined set of groups. Boot-Err assesses how much error score variance varies across contexts and whether that variation is substantially meaningful, a quantitative question about the magnitude of differences in random measurement error.

| Method | Target statistic | Inference | Main limitation |
|------------------|------------------|------------------|------------------|
| Reliability Generalization Meta-Analysis | Reliability coefficient | Ratio-based, differences in proportion of true to total score variance | No conclusion on magnitude of random measurement error |
| Residual invariance testing | Residual, item-level error variance | Model-specific, differences in magnitude of random measurement error | No conclusion on extent of differences in random measurement error |
| Boot-Err | Error score variance component | Extent of differences in error score variance | No established benchmarks for interpretation |

: Summary: comparison of approaches to heterogeneity in measurement quality {#tbl-summary}

Accessible methodological guidance exists for Reliability Generalization Meta-Analysis: first and foremost Rodriguez & Maeda [-@rodriguez2006], for additional information see Sánchez-Meca et al. [-@sánchez-meca2012], Henchy [-@henchy2013], and López-Ibáñez et al. [-@lopez2024]. Similarly, for measurement invariance testing literature is available: we firstly recommend reading Dimitrov [-@dimitrov2010] and for interested readers Steenkamp & Baumgartner [-@steenkamp1998], Van de Schoot et al. [-@vandeschoot2012] or Vandenberg & Lance [-@vandenberg2000] are available.

Boot-Err has been developed in the context of the increasingly common situation in which a scale is administered across a large number of independent contexts, such as the Many Labs projects [e.g. @ManyLabs1], Registered Replication Reports [e.g. @RRRMazar] or research stemming from the psychological science accelerator [e.g. @psychologicalscienceacceleratorself-determinationtheorycollaboration2022]. The remainder of this text therefore provides a step-by-step tutorial for Boot-Err, illustrated with data from a multi-lab replication project.

## Tutorial

The following section provides a hands-on guide to applying Boot-Err in practice. The goal is to compare error score variance of the same psychological scale across multiple independent administrations, and to test and quantify heterogeneity in random measurement error.

For illustration, we analyse the Conscientiousness subscale of the HEXACO-personality inventory [@lee2018] as administered in the multi-lab replication project reported by Verschuere et al. [-@RRRMazar]. In this project, Experiment 1 from Mazar et al. [-@mazar2008] was replicated across 19 independent laboratories. Each lab administered the same materials, including the HEXACO-inventory, to different participant samples scattered across the globe. This data-set therefore provides an ideal setting for Boot-Err: the same scale was administered across multiple independent samples under comparable but not identical conditions.

Importantly, our interest lies not in the experimental manipulation itself, but in the measurement properties of the HEXACO-Conscientiousness scale across laboratories. The HEXACO personality inventory was administered as a filler instrument between experimental tasks. Accordingly, scores of Conscientiousness should not be affected by the experimental manipulation. Each lab constitutes a separate administration, allowing us to examine whether random measurement error varies across administrations.

### Tutorial structure

This tutorial provides a step-by-step guide on how to investigate differences in measurement quality across administrations using Boot-Err. All analyses are conducted in R. The tutorial includes executable chunks that can be run directly on the reader's own machine by copying the code into R. A quarto-file for this tutorial, containing text and the executable code chunks can alternatively be found on OSF (osf.io/fkhpc) and GitHub (github.com/lbeinhauer/tutorial-boot-err). No additional proprietary software or supplementary materials are required. The data used within this tutorial is publicly available on the OSF (osf.io/fwnc2). We make use of an R-package to download and read the data-set. The entire procedure can be reproduced using the code presented in this section.

The tutorial is structured across several consecutive steps: (0) data preparation, (1) estimating score reliability, (2) estimating error score variance, (3) using bootstrapping to generate standard errors, (4) meta-analysis of error score variance, (5) backtransforming estimates, (bonus) relating error score variance to total score variance.

At the end of this tutorial, readers will be able to

-   estimate error score variance for each administration,

-   approximate the sampling variability of transformed error score variance using bootstrapping,

-   test whether error score variance differs across administrations, and

-   evaluate the magnitude of these differences in substantive terms

### Step 0 - Preparing Data

Before applying Boot-Err, the data must be prepared for analysis. Because data structures vary across projects, we walk through our own preparation here as a worked example. This procedure can be adapted to the requirements of the data at hand.

The following code chunk manages the packages required for this tutorial. By executing the code below, the user checks whether the required packages magrittr [@bache2022], dplyr [@wickham2014], osfr [@wolen2020], here [@wickham2023], zip [@Csardi2025], data.table [@dowle2019], metafor [@viechtbauer2015], boot [@canty2017] are already installed. If the packages are not yet installed, they will be downloaded and installed from the default CRAN-repository. All required packages are subsequently loaded.

```{r, output = F, warning = F}
# relevant libraries required for this script
packages <- c("magrittr", "dplyr", "osfr", "here", "zip", "data.table", "metafor", "boot")

# check, whether library already installed or not - install and load as needed:
apply(as.matrix(packages), MARGIN = 1, FUN = function(x) {
  
  # check if package is installed on system
  pkg_avail <- nzchar(system.file(package = x))   
  
   # load the package, if already installed
  if(pkg_avail){
    require(x, character.only = TRUE)            
    
  # install the package, if missing
  }else{
    install.packages(x)      
    
    # load after installation
    require(x, character.only = TRUE)             
  }
})
```

After loading the required packages, we can turn towards downloading the data-set from the OSF-repository. While the data cleaning and manipulation procedure is bound to be different for every research project, we briefly discuss how we come to a final data-set for this tutorial. Firstly, we download the zipped replication data-set from the openly available OSF-repository (osf.io/fwnc2), using the osfr R package. After unzipping, we can read the data-file in R.

```{r, warning = F, eval = F}
# create relevant directories, so data can be downloaded and stored
dir.create(here("Data"), showWarnings = FALSE)
dir.create(here("Data/Downloaded Data"), showWarnings = FALSE)


# use osfr-package to download RRR10 data on the Mazar et al. / Srull et al. replications
osfr::osf_retrieve_file("https://osf.io/fwnc2") %>% 
  osfr::osf_download(path = here("Data/Downloaded Data/"), conflicts = "overwrite")
# zip-file is stored at Downloaded Data

# unzip the file and move it to the corresponding directory
unzip(here("Data/Downloaded Data/Meta-Analysis_2018-07-09.zip"),
      files = "Meta-Analysis/Results/raw_data_corrected_MAA.csv",
      exdir = here("Data/Original Data/RRR10"),
      junkpaths = T)
```

For the purpose of this tutorial, we want to restrict our analyses to only those participants, which were included in the replicators' primary analyses. Accordingly, we end up with 4,674 participants from 19 laboratories. We achieve this in substep *a)* by identifying the relevant laboratories from the published manuscript and filtering out all laboratories that do not belong to the primary replication.

```{r, warning = F}
# HEXACO Conscientiousness

# a) Load data and select laboratories part of primary replication

# Thankfully, for the RRR10 project, the data can be taken from a file directly, without the need of combination:
pc_df <- as.data.frame(fread(here("Data/Original Data/RRR10/raw_data_corrected_MAA.csv")))

# retain observations only for participants eligible for analysis
pc_df <- pc_df[which(pc_df$inclusion == "inclusion both RRR" | pc_df$inclusion == "inclusion Mazar only"),] %>%
  mutate(source = lab.name)

# identify relevant labs for analysis
labs_in_paper <- c("Laine", "klein Selle & Rozmann", "Aczel", "Ferreira-Santos", "Meijer", "Loschelder", "Wick", "Suchotzki", 
                   "Sutan", "Vanpaemel", "Verschuere", "Wiggins", "Gonzalez-Iraizoz", "Koppel", "Birt", "McCarthy", "Evans", 
                   "Holzmeister", "Ozdogru")
labs_in_data <- unique(pc_df$source)
labs_in_data[8] <- "Gonzalez-Iraizoz"
labs_in_data[16] <- "Ozdogru"

# remove labs from data, which we do not need for analysis
labs_excl <- labs_in_data[!labs_in_data %in% labs_in_paper]
pc_df <- pc_df[which(!pc_df$source %in% labs_excl),]
```

The data-set contains a large amount of information not required for this analysis. Therefore, in substep *b)*, we retain only those columns from the data-set that contain information about the participants' responses to the HEXACO-items and in which laboratory the participant was assessed. Since some of the HEXACO-items are reverse-coded, we need to re-code half of the items, so the scale is identical across all items.

```{r, warning = F}
# b) Select relevant variables and re-code items that are coded in the opposite direction

# retain only those columns, which are related to the HEXACO-personality inventory
# we can do so by only retaining columns, whose names start with "hex"
# As we need to differentiate between scores from laboratories, we keep column "source"
pc_df <- pc_df[,c(which(names(pc_df) %in% c("source")),
                  grep("^hex", names(pc_df)))]

# recoding the hexaco items that need recoding
for(i in grep("^hex", names(pc_df))){
  pc_df[,i] <- as.integer(pc_df[,i])
}

# these are the numbers of the items, that need recoding
items_hex_recode <- c(30, 12, 60, 42, 24, 28, 53, 35, 41, 59, 28, 52, 10, 46, 9, 15, 57, 21, 26, 32, 14, 20, 44, 56, 1, 31, 49, 19, 55, 48)

# combining "hex" and number gives the column names
names_items_hex_recode <- paste0("hex", items_hex_recode) 

# adding _R for names of items, that are recoded
names_items_hex_recode_R <- paste0(names_items_hex_recode, "_R") 

# replace the original (un-recoded) scores with the recoded ones
pc_df[,names_items_hex_recode_R] <- 6 - pc_df[,names_items_hex_recode] 
```

Lastly, in substep *c)*, we remove all columns that contain responses on the other HEXACO-dimensions, which are not Conscientiousness. As a result, we have constructed a data-set, containing the responses on Conscientiousness from all participants that are part of the primary replication sample in Verschuere et al. [-@RRRMazar].

```{r, warning = FALSE}
# c) select items belonging to Conscientiousness subscale

# Conscientiousness
# items in Conscientiousness subscale
items_hex_CO <- c(2, 26, 8, 32, 14, 38, 50, 20, 44, 56) 

# did item need recoding?
names_items_hex_CO <- ifelse(items_hex_CO %in% items_hex_recode, paste0("hex", items_hex_CO, "_R"), paste0("hex",items_hex_CO)) 

# retain only columns concerning Conscientiousness and source
pc_df <- pc_df[,c(names_items_hex_CO, "source")]

# identify individual laboratories
labs <- unique(pc_df$source)
```

However, we find that some participants' responses on the Conscientiousness-items are missing in this data-set. While this was of no concern to the replication-authors, as the HEXACO-inventory served as a filler instrument, the same is not true for this tutorial. We will use Cronbach's alpha to estimate score reliability. Cronbach's alpha can not deal with missing responses. To keep it simple for this tutorial, we make use of list-wise deletion, where data for any participant with at least one missing item response is removed from the dataset. Alternative approaches attempting to impute the missing responses while retaining the general covariance-structure of the data-set are available. Great examples include Multiple Imputation [@schafer1999; @rubin2018] or Multivariate Imputation by Chained Equations [@vanbuuren2011]. Single value imputation methods like Mean or Mode Imputation restrict the data-set's covariance structure, leading to estimation issues and biases in subsequent analyses [@rubin2018].

In the following code chunk, we compute how many scores are missing for each individual item. Subsequently, we retain only scores from participants, where no single item score for Conscientiousness is missing.

```{r, warning = FALSE}
# d) dealing with missing responses

# compute number of missing values per item
colSums(is.na(pc_df))

# retain only complete observation sets
pc_df <- na.omit(pc_df)
```

Finishing step 0, we have constructed a data-set, containing only the complete response sets of participants belonging to the replication authors' primary analysis sample. Subsequently, we can begin to follow the Boot-Err procedure.

### Step 1 - estimating score reliability

The first step in the Boot-Err procedure aims to generate estimates of score reliability for each administration respectively. In the following code, we use *sapply()* to compute Cronbach's alpha separately for each laboratory in the data-set. Cronbach's alpha is generally defined as outlined in the following [@eq-alpha_formal; @cronbach1951].

$$\alpha=\frac{k^2\ \bar{\sigma}_{ij}}{\sigma^2_x}$$ {#eq-alpha_formal}

Here, $k$ corresponds to the number of items in the scale, $\bar{\sigma}_{ij}$ constitutes the average covariance between items and $\sigma^2_x$ constitutes the total score variance in the total test results.

However, computing the average item covariance can be avoided, computing Cronbach's alpha directly from the covariance matrix as in @eq-alpha_fast.

$$\alpha = \frac{k}{k-1}\left(1-\frac{\sum_{i=1}^k\sigma^2_{i}}{\sum_{i=1}^k\sum_{j=1}^k\ \sigma_{ij}}\right) $$ {#eq-alpha_fast}

Here, $\sigma^2_{i}$ describes the variance of scores on item $i$, while $\sigma_{ij}$ describes the covariance between items $i$ and $j$. @eq-alpha_fast generates the same estimate of Cronbach's alpha as in @eq-alpha_formal, but more easily implemented. Packages to compute Cronbach's alpha are available (e.g. psych, something else), but are more computationally demanding as several additional coefficients are returned. By using *sapply* to quickly compute across all laboratories, we can efficiently generate a vector containing estimates of Cronbach's alpha for each administration respectively.

```{r, warning = F}
# compute Cronbach's alpha across all administrations

# we use sapply as it tends to be faster than a for-loop
alpha_vec <- sapply(labs, FUN = function(lab){
  
  # select data for each administration separately
  dat <- pc_df %>% 
    filter(source == lab) %>% 
    select(-"source")
  
  # compute covariance matrix
  C <- cov(dat)
  
  # extract the number of items
  k <- dim(C)[1]
  
  # compute Cronbach's alpha efficiently
  alpha <- (k/(k - 1)) * (1 - sum(diag(C))/sum(C)) 
  
  return(alpha)
})

# return the estimate vector (rounded to 2 decimals)
round(alpha_vec, 2)
```

### Step 2 - estimating error score variance

After generating estimates of score reliability using Cronbach's alpha, we can turn towards Step 2 of the Boot-Err procedure - estimating error score variance. In @eq-errorvar we established that error score variance can be constructed from a reliability coefficient and the total score variance. This means, provided we generate an estimate of total score variance, we can use @eq-errorvar to estimate error score variance, as we already have an estimate of score reliability in the form of Cronbach's alpha readily available.

In the following code chunk we, again, use an *sapply*-function to efficiently repeat the same computation for each laboratory. We estimate the total score variance $\sigma^2_X$ by computing Conscientiousness scores as the average item score for each participant (generating the test score) and computing the variance of these test scores. Subsequently, we can easily use @eq-errorvar to generate a vector containing error score variance $\sigma^2_E$-estimates for each laboratory.

```{r}
# estimate error score variance for each administration

# we use sapply as it tends to be faster than a for-loop
var_X_vec <- sapply(labs, FUN = function(lab){
  
  # select data for each administration separately
  dat <- pc_df %>% 
    filter(source == lab) %>% 
    select(-"source")
  
  # compute the total score variance from individual test scores
  var_X <- dat %>% 
    rowMeans(.) %>% 
    var(.)
  
  return(var_X)
  
})

# compute error score variance from the estimates of reliability and total score variance
var_E_vec <-  (1-alpha_vec) * var_X_vec

# return the estimated vector (rounded to 3 decimals)
round(var_E_vec, 3)

```

### Step 3 - construct Standard Errors using Bootstrapping

As described in the introductory *Boot-Err*-section, we use a random-effects meta-analysis to estimate and test for heterogeneity in error score variances. Since estimation of total score variance and its components is not variance-stable, we need to use a variance-stabilizing transformation. Bartlett and Kendall [-@bartlett1946] show that using the natural log, as in @eq-stable_var, is a suitable variance-stabilizing transformation for total score variance estimates.

$$ T[\hat{\sigma}^2_x] = ln \left(\hat{\sigma}^2_x \right) $$ {#eq-stable_var}

While little evidence on variance-stabilizing transformations for score variance components is available, we have used the natural log for both the total and the error score variance component in our proposal for the Boot-Err [@beinhauer2025]. In the simulation scheme, we find a small but negligible bias in the estimates. We concluded that the transformation described in @eq-stable_var is suitable for the variance component of error score variance $\sigma^2_e$. Therefore, we are confident in our claim that @eq-stable_varE describes a suitable variance-stabilizing transformation.

$$ T[\hat{\sigma}^2_e] = ln \left(\hat{\sigma}^2_e \right) $$ {#eq-stable_varE}

However, in order to run a random-effects meta-analysis, we not only need a variance-stable estimate, but also a Standard Error, quantifying our uncertainty in that estimate. While we are aware of no analytic approximation for the Standard Error of log-transformed error score variance (natural log), we can use Bootstrapping techniques to generate robust Standard Error estimates for $T[\hat{\sigma}^2_e]$.

The package *boot* provides easy-to-use and efficient functions (including parallelization if desired) for Bootstrapping techniques. During the conception of this tutorial, we made use of version 1.3-31. In order to use *boot*, we need to provide the package with a user-written function, which needs to do two things: i) allow the *boot*-package to sample from our data-set (we do so by `dat_boot <- data[indices,]`), and ii) compute the parameter on the re-sampled data-set that we want to boot-strap. In the following code-chunk, we show a function that does just that.

```{r}
# define the user-written function to compute transformed error score variance

ln_var_E_boot_func <- function(data, indices){
  
  # essential for any boot-function: allow boot to sample from dataset rows
  dat_boot <- data[indices,]
  
  # we have to compute Cronbach's alpha again for each bootstrapped sample
  
  # compute covariance matrix
  C <- cov(dat_boot)
  
  # extract number of items
  k <- dim(C)[1]
  
  # compute Cronbach's alpha
  alpha <- (1 - sum(diag(C))/sum(C)) * (k/(k - 1))
  
  # we also have to compute error score variance again for each bootstrapped sample
  
  # compute total score variance
  var_X <- dat_boot %>% 
    rowMeans(.) %>% 
    var(.)
  
  # compute error score variance
  var_E <- var_X * (1 - alpha)
  
  # log-transform as variance stabilising transformation
  ln_var_E <- log(var_E)
  
  return(ln_var_E)
  
}
```

Now that we have a function available for *boot* to re-sample our data and estimate $T[\hat{\sigma}^2_e]$ in the bootstrapped samples, we need to supply our user-defined function to *boot* and compute the Standard Error from the bootstrapped estimates. Essentially, what we do here is estimate the Standard Error of log-transformed error score variance, by computing the standard deviation of all bootstrapped estimates for a single laboratory, see @eq-bootSE.

$$ SE\left[T[\hat{\sigma}^2_e]\right] = \sqrt{\frac{1}{n}\sum_{i=1}^n(T[\hat{\sigma}^2_e]_i - \bar{T}[\hat{\sigma}^2_e])^2}$$ {#eq-bootSE}

Here, $n$ describes the number of bootstrap-samples drawn, whereas $\bar{T}[\hat{\sigma}^2_e]$ describes the mean of the log-transformed error score variance estimates of the $n$ bootstrap samples. In the following code-chunk, we use the *boot* package to estimate $SE\left[T[\hat{\sigma}^2_e]\right]$ from 3000 bootstrap samples for each laboratory respectively. Again, we make use of an *sapply*-function to efficiently compute across all laboratories.

```{r, warning = FALSE}
# approximate the standard error for error score variance

ln_var_E_SE_vec <- sapply(labs, FUN = function(lab){
  
  # select data for each administration separately
  dat <- pc_df %>% 
    filter(source == lab) %>% 
    select(-"source")
  
  # use boot and our user-defined function to efficiently generate transformed error score variance in 3000 bootstrapped samples
  boot_estimates <- boot::boot(dat, 
                               statistic = ln_var_E_boot_func,
                               R = 3000)
  
  # compute the standard error
  SE <- sd(boot_estimates$t)
  
  return(SE)
  
})
```

Lastly, we still need to formally transform and store the individual estimates of $T[\hat{\sigma}^2_e]$ for each laboratory (so far we only did this for the bootstrap-samples). However, at this point, computing this transformation is trivial. To make things easier for us in the following steps, we combine the estimates of log-transformed error score variance and its Standard Errors here in a single data-frame. We add the laboratory names to the data-frame, so estimates can be linked to their respective sources.

```{r, warning = FALSE}
# combine standard errors and estimates in a data-frame

# transform individual estimates of error score variance
ln_var_E_vec <- log(var_E_vec)

# combine estimates, standard error and lab-name in data-frame
ln_var_E_df <- data.frame(ln_var_E = ln_var_E_vec, 
                          ln_var_E_SE = ln_var_E_SE_vec, 
                          laboratory = labs, 
                          row.names = NULL)

# return the estimates and round the numeric variables (ln-estimate and SE)
ln_var_E_df %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

Concluding step 3, we have gathered all estimates we need, to perform a random-effects meta-analysis on (log-transformed) error score variances.

### Step 4 - meta-analysis of error score variance

For this step, we estimate the heterogeneity of log-transformed error score variance and test whether it can be statistically distinguished from zero. We use the R-package *metafor*, as it provides an efficient and easy-to-use implementation of random-effects meta-analyses for any parameter [@metafor]. Using the *rma*-function with the argument *method = "REML"* ensures that we use the REML-estimator of heterogeneity. Hönekopp and Linden [-@hoenekopp2022] have shown that the REML-estimator tends to outperform other estimators across a wide range of conditions.

The following chunk will run a random-effects meta-analysis in *metafor*, generating an estimate of the meta-analytic mean (in the output `estimate`), estimates of absolute heterogeneity in standard deviation and variance (in the output `tau` and `tau^2`) and estimates of relative heterogeneity in the form of $I^2$ and $H^2$ (in the output `I^2` and `H^2`). Additionally, *metafor* provides the results for two hypothesis tests. Firstly, the test for heterogeneity (`Test for Heterogeneity` in the output), assessing whether the estimate of heterogeneity $\tau^2$ can be statistically distinguished from $0$. Secondly, the test for the meta-analytic effect size, assessing whether the meta-analytic aggregate can be statistically distinguished from $0$ (in the Output `Model results`).

```{r}
# perform the random-effects meta-analysis on log-transformed error score variance
ln_var_E_rma <- metafor::rma(data = ln_var_E_df,
                             yi = ln_var_E,
                             sei = ln_var_E_SE,
                             method = "REML")

ln_var_E_rma
```

The significance tests indicate two things for the log-transformed error score variance of Conscientiousness across replications in Verschuere et al. [-@RRRMazar]: Firstly, we find that the meta-analytic estimate of `r paste0("$\\mu_{T[\\hat{\\sigma}^2_e]} = ", round(ln_var_E_rma$b[1], 3), "$")` is significantly different from zero (`r paste0("$z = ", round(ln_var_E_rma$zval, 3), "$")`, `r paste0("$p ", ifelse(ln_var_E_rma$pval < .001, yes = paste0('< .001'), no = paste0('= ', round(ln_var_E_rma$pval, 3))), "$")`), using a significance level of 5%. However, this information is not particularly meaningful, the estimate is on the log-transformed scale and interpretation is not straightforward. Secondly and of greater relevance here, we find that the test for heterogeneity indicates that there is statistically significant heterogeneity in log-transformed error score variances (`r paste0("$Q(", ln_var_E_rma$k - 1, ") = ", round(ln_var_E_rma$QE, 3), "$")`, `r paste0("$p ", ifelse(ln_var_E_rma$QEp < .001, yes = paste0('< .001'), no = paste0('= ', round(ln_var_E_rma$QEp, 3))), "$")`). Even though the tested estimates are still on the log-scale, we can nonetheless infer from this result, that error score variance $\sigma^2_E$ is not identical across all administrations of the Conscientiousness-scale.

However, the estimate of `r paste0("$\\tau^2_{T[\\hat{\\sigma}^2_e]} = ", round(ln_var_E_rma$tau2, 3), "$")` can not be easily interpreted due to the log-transformation. Instead, the estimates of relative heterogeneity may be interpreted despite the transformation. The estimate of `r paste0("$I^2 = ", round(ln_var_E_rma$I2, 2), "$")` implies that `r round(ln_var_E_rma$I2, 2)`% of the variance observed in the estimate of error score variance are due to actual heterogeneity in the population, not due to sampling error. Similarly, the estimate of `r paste0("$H^2 = ", round(ln_var_E_rma$H2, 2), "$")` implies that the heterogeneity is `r round(ln_var_E_rma$H2, 2)` times as large as the variance induced by sampling error. According to standard conventions [@higgins2002], both estimates imply a strong or large extent of heterogeneity. However, both estimates of relative heterogeneity are often criticized, as they depend strongly on sample size and are easily manipulated [@borenstein2017].

Instead, we should turn towards an informative interpretation of the absolute heterogeneity in terms of $\tau^2$ or $\tau$. To do so, we need to back-transform our newly generated estimates to the original scale. We do so in Step 5.

### Step 5 - Back-transforming estimates

The transformation of score variance (components) using the natural log not only serves to generate variance-stable estimates. Additionally, it comes with the added benefit, that we can assume $T[\sigma^2_e]$ to follow a normal distribution. Viechtbauer and López-López [-@viechtbauer2022] model variance components using a log-link, Raudenbush & Bryk [-@raudenbush1987] also make use of log-transformations to model standard deviations. We follow their approaches, assuming that score variance components can be described using a log-normal distribution. This means that the log-transformed estimates accordingly follow a normal distribution. This is very helpful for the application of the Boot-Err technique, as one central assumption of a random-effects meta-analysis is that the parameter follows a normal distribution. Additionally, it helps us back-transform the estimates of meta-analytic mean $\mu_{T[\sigma^2_e]}$ and its heterogeneity $\tau^2_{T[\sigma^2_e]}$, as equations are readily available.

$$ \hat{\mu}_{\sigma^2_e} = \exp\left(\hat{\mu}_{T[\sigma^2_e]} + \frac{1}{2} \hat{\tau}^2_{T[\sigma^2_e]}\right) $$ {#eq-backtransf_mu}

@eq-backtransf_mu demonstrates how the mean error score variance $\hat{\mu}_{\sigma^2_e}$ can be estimated from the meta-analytic estimates of $T[\sigma^2_e]$. Here, $\hat{\mu}_{T[\sigma^2_e]}$ represents the estimate of meta-analytic mean of log-transformed error score variance and $\hat{\tau}^2_{T[\sigma^2_e]}$ describes the estimate of its heterogeneity. Similarly; @eq-backtransf_var can be used to derive the heterogeneity of error score variance $\hat{\tau}^2_{\sigma^2_e}$ on its original scale, from the same meta-analytic estimates of $T[\sigma^2_e]$.

$$ \hat{\tau}^2_{\sigma^2_e} = \exp\left(\hat{\tau}^2_{T[\sigma^2_e]} - 1\right) \cdot \exp\left(2 \hat{\mu}_{T[\sigma^2_e]} + \hat{\tau}^2_{T[\sigma^2_e]}\right) $$ {#eq-backtransf_var}

In the following chunk, we define two functions which compute [Equations -@eq-backtransf_mu] and [-@eq-backtransf_var] from the *metafor*-output. The functions only take the *metafor*-object generated from the *rma*-function as input and return the estimates $\hat{\mu}_{\sigma^2_e}$ and $\hat{\tau}^2_{\sigma^2_e}$ respectively.

```{r}
# two functions to back-transform heterogeneity and meta-analytic estimate

# back-transform meta-analytic estimate (the weighted mean)
backtransform_var_mu <- function(rma_obj){
  exp(rma_obj$b[1] + (.5*rma_obj$tau2))
}

# back-transforming meta-analytic heterogeneity
backtransform_var_tau2 <- function(rma_obj){
  (exp(rma_obj$tau2) - 1) * exp((2 * rma_obj$b[1]) + rma_obj$tau2)
}


round(backtransform_var_mu(ln_var_E_rma), 3)

round(backtransform_var_tau2(ln_var_E_rma) %>% 
  sqrt(.), 3)
```

Running the two functions in the chunk above returns two estimates: The meta-analytic mean error score variance of the HEXACO-Conscientiousness scale across 19 laboratories is $\hat{\mu}_{\sigma^2_e} = `{r} round(backtransform_var_mu(ln_var_E_rma), 3)`$; The estimate for its heterogeneity is $\hat{\tau}_{\sigma^2_e} =`{r} round((backtransform_var_tau2(ln_var_E_rma) %>%   sqrt(.)), 3)`$. To understand what this implies, we can assess the relative size of heterogeneity, also known as the coefficient of variation, which is computed as $CV = \frac{\tau}{\mu}$. Here we find a CV of $CV_{\sigma^2_e} = `{r} round((backtransform_var_tau2(ln_var_E_rma) %>% sqrt(.))/backtransform_var_mu(ln_var_E_rma), 3)`$. This means that the extent of the differences in error score variances is approximately `{r} round((backtransform_var_tau2(ln_var_E_rma) %>% sqrt(.))/backtransform_var_mu(ln_var_E_rma), 3) * 100`% of the mean error score variance found across these laboratories. Based on the random-effects model, true error score variance across administrations can be expected to lie approximately between `r round(exp(ln_var_E_rma$b[1] - sqrt(ln_var_E_rma$tau2)), 3)` and `r round(exp(ln_var_E_rma$b[1] + sqrt(ln_var_E_rma$tau2)), 3)`. These values can be obtained by adding and subtracting the estimated heterogeneity $\hat{\tau}_{\sigma^2_e}$ from the meta-analytic mean $\hat{\mu}_{\sigma^2_e}$ (on the transformed scale) and subsequently back-transforming to the original variance metric. 

As the substantive interpretation of the extent of heterogeneity in error score variance requires domain knowledge about expected levels of measurement error for this scale, we cannot evaluate directly whether the observed heterogeneity is practically meaningful. However, we can construct additional parameters to gain a deeper understanding. In the following Bonus-Step, we delve deeper into the insights generated by the Boot-Err technique.

### Bonus-Step - Relating error score variance to total score variance

Heterogeneity in total score variance is often interpreted as if it reflects differences in measurement precision. The Bonus-Step tests this directly by asking how much of the observed heterogeneity in total score variance can actually be attributed to error score variance, and how much must be due to genuine differences in true scores.

Score reliability $\rho_{xx'}$ is defined as the relative extent to which total score variance observed is attributable to actual differences in the true scores, not to measurement imprecision, see @eq-reliability. However, that also means that $1 - \rho_{xx'}$ describes the relative extent to which total score variance is attributable to random measurement error.

Similarly, on the meta-analytic scale, we can look at the heterogeneity in total score variance $\sigma^2_x$, and attempt to understand how much of its heterogeneity can be attributed to differences in error score variance. Substantial parts of said heterogeneity might also be better explained by differences in the true score variance $\sigma^2_t$. Fortunately, if we follow Steps 3 through 5 for the total score variance $\sigma^2_x$ instead, we can assess to what extent its heterogeneity can be attributed to differences in true score variance $\sigma^2_t$ or measurement error $\sigma^2_e$.

In the following chunk, we briefly run through these steps to estimate heterogeneity in total score variance $\sigma^2_x$.

```{r}
# perform steps 3 through 5 for total score variance

# define the function for boot to bootstrap transformed total score variance
ln_var_X_boot_func <- function(data, indices){
  
  # allow boot to resample from data-set
  dat_boot <- data[indices,]
  
  # compute total score variance
  var_X <- dat_boot %>% 
    rowMeans(.) %>% 
    var(.)
  
  # transform total score variance
  ln_var_X <- log(var_X)
  
  return(ln_var_X)
}


# use Bootstrapping to compute standard error of log-transformed total score variance
ln_var_X_SE_vec <- sapply(labs, FUN = function(lab){
  
  # select data only for specific administrations
  dat <- pc_df %>% 
    filter(source == lab) %>% 
    select(-"source")
  
  # use boot to bootstrap log-transformed total score variance
  boot_estimates <- boot::boot(dat, 
                               statistic = ln_var_X_boot_func,
                               R = 3000)
  
  # estimate standard error
  SE <- sd(boot_estimates$t)
  
  return(SE)
  
})

# transform total score variance estimates
ln_var_X_vec <- log(var_X_vec)

# combine estimates of total score variance and its standard error in single df
ln_var_X_df <- data.frame(ln_var_X = ln_var_X_vec, 
                          ln_var_X_SE = ln_var_X_SE_vec, 
                          laboratory = labs, 
                          row.names = NULL)

# run random-effects meta-analysis
ln_var_X_rma <- metafor::rma(data = ln_var_X_df,
                             yi = ln_var_X,
                             sei = ln_var_X_SE,
                             method = "REML")

# round back-transformed meta-analytic estimate to 3rd decimal
round(backtransform_var_mu(ln_var_X_rma), 3)

# round back-transformed heterogeneity estimate to 3rd decimal
round(backtransform_var_tau2(ln_var_X_rma) %>% 
  sqrt(.), 3)

```

We find a meta-analytic mean of total score variance $\hat{\mu}_{\sigma^2_x} = `{r} round(backtransform_var_mu(ln_var_X_rma), 3)`$ and a heterogeneity of $\hat{\tau}_{\sigma^2_x} = `{r} round((backtransform_var_tau2(ln_var_X_rma) %>% sqrt(.)), 3)`$. Under CTT, total score variance is defined as the sum of true and error score variance $\sigma^2_x = \sigma^2_e + \sigma^2_t$, which are assumed to be independent. This means that we can assess how much heterogeneity in total score variance can actually be explained by differences in measurement error:

$$ R = \frac{\hat{\tau}^2_{\sigma^2_e}}{\hat{\tau}^2_{\sigma^2_x}} $$ {#eq-ratio}

In @eq-ratio, we define the ratio-variable $R$ as the ratio of $\hat{\tau}_{\sigma^2_e}$ to $\hat{\tau}_{\sigma^2_X}$. For Conscientiousness, we find a ratio of $R  = `{r} round((backtransform_var_tau2(ln_var_E_rma)) / (backtransform_var_tau2(ln_var_X_rma)), 3)`$. This means that about `{r} round((backtransform_var_tau2(ln_var_E_rma)) / (backtransform_var_tau2(ln_var_X_rma)), 3) * 100`% of heterogeneity in $\sigma^2_x$ could be explained by differences in random measurement error. This also means that `{r} round((1 - (backtransform_var_tau2(ln_var_E_rma)) / (backtransform_var_tau2(ln_var_X_rma))), 3) * 100`% of this heterogeneity is explained by differences in true score variance $\sigma^2_t$.

By following the steps from 1 through 5 we have assessed differences in measurement quality of Conscientiousness across different administrations. In doing so, including the Bonus-Step, we have generated a number of estimates and parameters that help us understand to what extent these differences are substantive. In the following section, we turn towards a substantive interpretation of what these results imply for the use of the HEXACO-Conscientiousness inventory.

### Conclusion

Boot-Err essentially consists of 5 steps, where we estimate score reliability (step 1), estimate error score variance (step 2), use appropriate variance-stabilizing transformations paired with Bootstrapping to estimate standard errors (step 3), perform meta-analyses of error score variance (step 4) and back-transform estimates to the original scale (step 5). In the previous section, we provided custom functions in R for researchers to perform a Boot-Err analysis on data obtained from a single measuring instrument administered across different contexts. Combining the individual steps allows for an assessment of whether there is heterogeneity in random measurement error across administrations of the same scale.

Having worked through these steps, readers are now equipped to estimate error score variance for each administration, approximate its sampling variability using bootstrapping, test whether error score variance differs significantly across administrations, and evaluate the magnitude of those differences in substantive terms, corresponding to the four learning objectives set out at the start of the tutorial.

Making use of the data supplied by Verschuere et al. [-@RRRMazar], we used Boot-Err to understand to what extent there are differences in absolute measurement precision across 19 laboratories where the HEXACO-Conscientiousness personality scale was administered. We not only find statistically significant differences in error score variance, but were also able to estimate the absolute and relative extent of these differences. Consequently, the evidence strongly suggests that the random measurement error in the Conscientiousness scores is not constant across samples. We find that the heterogeneity of `{r} round(backtransform_var_tau2(ln_var_E_rma), 3)` is about `{r} round((backtransform_var_tau2(ln_var_E_rma) %>% sqrt(.))/backtransform_var_mu(ln_var_E_rma), 3) * 100`% of the mean error score variance. Across administrations of the same scale, we can expect error score variance to vary at least between `r round(exp(ln_var_E_rma$b[1] - sqrt(ln_var_E_rma$tau2)), 3)` and `r round(exp(ln_var_E_rma$b[1] + sqrt(ln_var_E_rma$tau2)), 3)`.

In order to better understand what these differences mean, we additionally uncovered to what extent the heterogeneity in total score variance can be attributed to these differences in measurement error. Here, we found that the overwhelming majority of heterogeneity, `{r} round((1 - (backtransform_var_tau2(ln_var_E_rma)) / (backtransform_var_tau2(ln_var_X_rma))), 3) * 100`%, can not be explained by differences in measurement error. For the Conscientiousness-scale, this implies that, while there clearly are substantial differences in measurement error across groups, the distribution of true scores varies much more strongly across groups. Accordingly, differences in total score variance can not be attributed to differences in measurement error. Instead, Conscientiousness is dispersed highly differently across those groups. In Beinhauer et al. [-@beinhauer2026] we discuss potential causes and implications of this phenomenon.

## Discussion

Measurement precision and how strongly it diverges across administrations of the same instrument is a central question in social and behavioural sciences [e.g.; @shaw2020; @garcía2022; @botella2010]. Differences in random measurement error are not only interesting to researchers assessing the quality of a measuring instrument, but these differences also have clear implications concerning hypothesis testing and standardization of effect sizes [e.g.; @olsson-collentine2020; @wiernik2020].

The three methods discussed in this text each address this concern from a distinct angle. As summarised in Table @tbl-summary, these three methods are complements rather than competitors, as each answers a distinct question, and the choice should be driven by the research goal, not convention.

In the tutorial-section, we have demonstrated how Boot-Err can be applied in practice using the HEXACO-Conscientiousness scale across 19 laboratories. We found statistically significant heterogeneity in error score variance, and were able to estimate both its absolute magnitude and its contribution to total score variance heterogeneity. Importantly, the Bonus-Step revealed that the overwhelming majority of heterogeneity in total score variance could not be attributed to differences in measurement error. Instead, differences in true score variance were the dominant source of variation. This finding illustrates a broader point, as heterogeneity in observed score variance is frequently and perhaps reflexively attributed to measurement imprecision, when in fact differences in the construct itself across groups may be the more substantive explanation. Boot-Err makes it possible to formally distinguish between these two sources rather than assuming one.

Like any statistical method, Boot-Err rests on assumptions that should be made explicit. Total and error score variance should follow a log-normal distribution, and administrations should be independent rather than nested. Additionally, a number of assumptions relate to the choice of the reliability estimator. Here, we made use of Cronbach's alpha, which typically comes with assumptions of essential tau-equivalence and unidimensionality [@cronbach1951]. While these assumptions are widely criticized and rarely hold up in real data [@sijtsma2009; @mcneish2018; @mcdonald1999], we need to remind ourselves that these assumptions also relate to typical scoring methods of these measuring instruments. The HEXACO-personality score for each dimension is computed by taking the average of the relevant item scores [@lee2018]. Therefore, even stricter assumptions going beyond tau-equivalence and unidimensionality apply to the scoring mechanisms. Because the HEXACO-personality inventpry makes use of sum/average scoring, the tau-equivalence assumption is already embedded in the scoring procedure itself. Using a reliability estimator that imposes weaker assumptions would give a score reliability estimate inconsistent with how the scale is actually scored.

We hope this tutorial lowers the barrier for researchers who wish to go beyond asking *whether* measurement quality differs across contexts and begin asking *how much* it differs. The increasing prevalence of large-scale collaborative projects means that data well-suited for Boot-Err analyses are becoming less of an exception. Beyond measurement quality itself, future applications could use Boot-Err to examine whether heterogeneity in error score variance systematically covaries with study-level moderators, such as sample characteristics, administration mode or cultural context. More broadly, we encourage researchers to treat measurement error not as a fixed property of a scale, but as a quantity that varies, can be estimated, and deserves as much attention as the constructs being measured.

## References

::: {#refs}
:::
