---
title: "Tutorial"
format: pdf
bibliography: bibliography/Boot-Err_Tutorial.bib
csl: bibliography/apa-6th-edition.csl
---

# Measuring measurement quality: A critical comparison of three approaches to random error

assessment

new structure should follow:

Some things sound alike. We explain what each individual thing does. We identify use cases for each individual thing. There are better tutorials and discussions out for RG-MA and MI. What's missing is a tutorial on Boot-Err. That follows here.

Example use cases:

RG-MA where is scale relatively more informative in one pop than the other e.g. ability test - does it discriminate better among uni than general public - students likely less variation among abilities than general public essentially: signal to noise-ratio of interest. If rel largely equal, test can be used in both contexts to differentiate between levels of ability

Residual Invariance Testing cross-cultural CFA of e.g. personality dimension, needs to know whether multi-factor scale meets statistical requirements for group comparison (spec.: does measurement model hold equivalently) Test is confirmatory: does model hold likely less groups

Boot-Err Same scale across larger number of labs / cultures, wants to know how much measurement quality varies - e.g. how much does it contribute to variability in effect sizes, is variance large enough to be relevant for pooling purposes

## Introduction

Researchers frequently administer the same psychological scale across different samples, settings or populations. A scale may be translated into another language, administered online instead of in person, or completed by participants with different educational or cultural backgrounds. In such situations, an important question arises: does the instrument produce scores of equivalent measurement quality across these administrations?

At first glance, this question appears straightforward and several established methods seem to address it. Reliability Generalization Meta-Analysis, residual invariance testing and Boot-Err all speak to the question of whether measurement quality differs across contexts. Because they share this surface-level concern, they might seem interchangeable. However, each method answers a subtly but importantly different question. Applying the wrong tool to a given research problem can lead to misleading conclusions.

The goal of this text is threefold: (1) to clarify what each method actually does, (2) to identify the research contexts each method is best suited for, and, because accessible tutorials already exist for Reliability Generalization Meta-Analysis and residual invariance testing, (3) provide a hands-on tutorial for Boot-Err, where such guidance is currently lacking.

Throughout this text, we focus on random measurement error as a central component of measurement quality. Random measurement error refers to unsystematic fluctuations in observed scores around the values they would take on in the absence of measurement imprecision. Unlike systematic error, which shifts observed scores in a particular direction, random error introduces general variation sometimes referred to as noise. On average, observed scores may still be accurate, but individual measurements vary unpredictably around their underlying true values.

If the same instrument is administered in two different contexts, differences in random measurement error imply differences in measurement quality. However, the three methods discussed here do not address this in the same way. Instead, they reflect different perspectives on what it means for measurement quality to be equivalent, and they are therefore suited to different research questions.

### Reliability generalization meta-analysis

The first, most intuitive and widely used approach is to compare reliability coefficients across administrations of the same scale. The conceptual idea behind reliability coefficients can be described in terms of classical test theory. Classical test theory implies that each score $X$ is made up of true score $T$ and error score $E$:

$$ X = T + E$$ {#eq-CTT}

Reliability coefficients, such as Cronbach's Alpha, McDonald's Omega or Guttmann's Lambda IV quantify the proportion of true score variance present in the scores:

$$\rho_{XX'} = \frac{\sigma^2_T}{\sigma^2_X}$$ {#eq-reliability}

where $\rho_{XX'}$ is score reliability, $\sigma^2_T$ is the true score variance and $\sigma^2_X$ is the total score variance, made up of true and error score variance ($\sigma^2_X = \sigma^2_T + \sigma^2_E$, where $\sigma^2_E$ is the error score variance).

This true score variance $\sigma^2_T$ describes the variation in the scores, that can be attributed to genuine differences in the underlying factor being measured. Error score variance $\sigma^2_E$ describes the variation in the scores, that can be attributed to random, unsystematic variation - random measurement error that is introduced by the instruments inability to perfectly measure the underlying factor. Reliability coefficients thereby express the proportion of observed score variance that cannot be attributed to random measurement error, and takes on values between 0 and 1.

Reliability Generalization Meta-Analysis extends this idea by formally testing whether observed differences in score reliability exceed what would be expected from sampling error alone. Because coefficients derived from samples are always prone to sampling variability, on can not simply compare point estimates across administrations. Meta-analytic techniques model this sampling error and test for heterogeneity. (Meta-analytic) heterogeneity refers to variance, that exceeds the variation that sampling fluctuation alone would produce. If significant heterogeneity is detected, this indicates that reliability differs across administrations in a way, that is likely not attributable to chance. The method additionally yields an estimate of the magnitude of this heterogeneity.

**What question does it answer?** Reliability Generalization Meta-Analysis addresses a ratio-based question: in which contexts is the scale relatively more informative, in the sense that a greater proportion of observed score variance reflects true differences among individuals?

**When is this the right tool?** This approach is well suited when the research question concerns the *signal-to-noise ratio* of a scale across populations that are expected to differ in the spread of the underlying construct. Consider an ability test administered both to university students and to the general public. Students likely show less variation in ability than the general population. It is (used to be?) common practice to validate such a test in a sample of university students, as these are readily available. However, if a researcher wishes to apply the ability test to the sample of the general population in order to differentiate between that population's ability levels, the absolute extent of random measurement error is not of interest.

Instead, the relative amount of measurement error, the signal-to-noise ratio as assessed through score reliability is of central interest in terms of measurement quality. If score reliability for the general population samples is approximately equal to score reliability for the university student sample, then the ability test discriminates between the levels of ability in both groups equally well. If it differs, the reasons for that differences and their practical implications warrant further investigation. Reliability Generalization Meta-Analysis can reveal this directly, informing decisions about whether the test usefully differentiates among individuals in both contexts.

**Limitations** Because score reliability is a ratio, its value is affected by differences in true score variance just as much as by differences in error score variance. Consider the following for our ability test example. Suppose we observe a Cronbach's alpha of .67 both for the student sample and the general population sample.

$$\rho_{xx'\ students} = \frac{2}{2 + 1} = .67 \qquad \rho_{xx'\ general} = \frac{3}{3 + 1.5} = .67$$ {#eq-RGMA2}

In this pattern, we observe identical score reliability and differences in true and error score variance. True and error score variance retain the same proportion in both samples. This means that the relative amount of measurement error in the observed scores is identical across both samples. This is the core mechanism, why the test can be used to differentiate equally well between different ability levels across both samples.

If, on the other hand, the test was administered to both samples with the goal to differentiate ability levels between the two samples, measurement quality can no longer be viewed as equivalent between the populations. The differences in error score variance ($1$ in the student sample, $1.5$ in the general population) imply, that individual measurements made in the student sample are more precise than those in the general population. Therein, the test does not discriminate equally well between university students' ability levels as for the general population's ability levels. The question regarding the test's measurement quality shifted from a signal-to-noise ratio perspective, to a perspective where the absolute magnitude of measurement error needs to be evaluated.

Consequently, Reliability Generalization Meta-Analysis answers a ratio-based question about measurement quality, not a question about differences in the absolute magnitude of measurement error.

### Measurement invariance testing

A second perspective is offered by measurement invariance testing, grounded in confirmatory factor analysis and typically implemented using structural equation modelling. Rather than summarising measurement quality as a single coefficient per sample, this framework models the relationship between latent factors and observed item responses and tests, whether this relationship holds equivalently across groups.

Measurement invariance testing proceeds through a series of increasingly constrained models, typically four levels:

-   Configural invariance: The same factor structure holds across all groups.

-   Metric invariance: Factor loadings are identical across groups.

-   Scalar invariance: Intercepts are identical, meaning systematic differences in observed scores across groups can be attributed to differences in latent factors.

-   Residual invariance: Residual variances describing variation in item scores not explained by the latent factor are identical across groups.

The last level is of greatest relevance here. In classical test theory terms, residual variance can be interpreted as item-level error score variance (with some caveats, see ref XY). The test for residual invariance evaluates whether these variances are equal across groups by comparing model fit between an unconstrained model (residual variances freely estimated) and a constrained model (residual variance held equal). A statistically significant $\Delta\chi^2\ (p < .05)$ suggests that the constraint meaningfully worsens model fit, indicating a lack of residual invariance. Because chi-square tests are sensitive to sample size, researchers typically supplement this with changes in approximate fit indices: small values (e.g. $\Delta CFI \leq .1$, $\Delta RMSEA \leq .015$, $\Delta SRMR \leq .01$) are taken as evidence, that residual invariance holds.

**What question does it answer?** Residual invariance testing addresses a *confirmatory, model-based* question: does the measurement model hold equivalently across groups, in the sense that item-level error variances are statistically indistinguishable?

**When is this the right tool?** This approach is well suited when the primary goal is to verify that the statistical requirements for group comparison are met. An common example is cross-cultural research on a personality dimension and needing to establish that scale scores are genuinely comparable. It is particularly appropriate when working with a smaller, well-defined set of groups and when the question is fundamentally confirmatory, whether the model holds. Returning to the example above (@eq-RMA), where score reliability coefficients were identical despite differing error score variances. For that example we established, that differences in ability may be meaningfully compared within groups, as we learned from Reliability Generalization Meta-Analysis. However, in order to ensure that ability levels may be compared across groups, we need to assess whether the magnitude in error score variance is equivalent across groups. Residual invariance testing, given adequate power, would correctly flag the difference in error score variance that reliability coefficients would have missed.

**Limitations** While residual invariance testing can detect whether error score variances differ, it does not quantify the magnitude of those differences. Whether the discrepancy in error score variance is .5 or 17, the test result is qualitatively the same, as invariance either does or does not hold. When the scale is administered across a large number of contexts, and the research question concerns not merely whether measurement quality varies, but how *much*, residual invariance testing is not sufficient to derive a meaningful answer.

### Boot-Err - inspecting differences in random measurement error

Boot-Err addresses this gap. Like residual invariance testing, it conceptualises measurement quality in terms of the absolute magnitude of measurement error rather than its total variance. Unlike residual invariance testing, it provides an estimate of the *magnitude of heterogeneity* in error score variance across administration contexts, along with a formal statistical test.

The key identity underlying Boot-Err is that error score variance can be derived from observed score variance and a reliability estimate:

$$\hat{\sigma}^2_E = (1 - \hat{\rho}_{xx'})\ \hat{\sigma}^2_X$$ {#eq-errorvar}

Bootstrapping is used to approximate the sampling variability of this estimate, expressed as a standard error. After applying appropriate variance-stabilising transformations, error score variance estimates from multiple administrations are entered into a random-effects meta-analysis (with subsequent back-transformation). The random-effects framework is essential here: unlike fixed-effects meta-analysis, which assumes zero heterogeneity in the population, random-effects models explicitly accommodate and estimate the between-administration variance that Boot-Err is designed to quantify. The result is both a test of whether differences in error score variance are statistically significant and an estimate of their magnitude, expressed as heterogeneity $\tau^2_{\sigma^2_E}$.

**What question does it answer?** Boot-Err addresses a **quantitative** question about measurement error: how much does random measurement error vary across administration contexts, and is that variation large enough to matter?

**When is this the right tool?** Boot-Err is best suited when the same scale has been administered across a large number of independent contexts, such as across laboratories in a multi-lab replication project, or across cultural groups in a large-scale cross-national study. The research question largely pertains the *magnitude* of variability in measurement quality. This is relevant, for example, when trying to understand how much variation in measurement error contributes to heterogeneity in effect sizes across labs, or when deciding whether differences in measurement error are large enough to complicate pooling of data across sites. Because Boot-Err operates on error score variance directly, rather than reliability ratios, it is not confounded by differences in true score variance across groups. Differences in random measurement error can still be detected and quantified.

Returning to the example in @eq-RMA2, where reliability coefficients were equal but error score variances differed ($\sigma^2_{E, students} = 1$, $\sigma^2_{E, general} = 1.5$). Boot-Err (if appropriately powered), would detect this difference and estimate the heterogeneity as $\tau^2_{\sigma^2_E} = .065$ ($\tau_{\sigma^2_E} = .25$). Given substantive knowledge about the scale and the expected magnitude of true and error score variance, researchers can evaluate whether this degree of variation has meaningful practical implications for group comparisons or effect sizes. In the chapter *Bonus-Step* of the following tutorial, we highlight perspectives on how such a question can be discussed and framed.

**Limitations** Identify some, to be fair.

### Summary and scope

The three methods thus address three distinct questions, summarised in table X. Reliability generalization meta-analysis assesses whether the *proportion* of true score variance differs across contexts, a ratio-based question about relative signal quality. Residual invariance testing assesses whether the measurement model holds equivalently, a confirmatory question about whether item-level error variances are statistically indistinguishable across a defined set of groups. Boot-Err assesses how much error score variance varies across contexts and whether that variation is substantially meaningful, a quantitative question about the magnitude of differences in random measurement error.

Accessible methodological guidance exists for reliability generalization meta-analysis (e.g. refs) and measurement invariance testing (e.g. refs). Boot-Err has been developed in the context of the increasingly common situation in which a scale is administered across a large number of independent contexts. The reminder of this text therefore provides a step-by-step tutorial for Boot-Err. We demonstrate how to estimate error score variance for each administration, approximate its sampling variability using Bootstrapping, and test and quantify heterogeneity in error score variance using random-effects meta-analysis, illustrated with data from a multi-lab replication project.

## Tutorial

Having demonstrated the conceptual foundations of Boot-Err, we now demonstrate how the method can be implemented in practice. The goal of this tutorial section is to estimate and compare error score variance of the same psychological scale across multiple independent administration contexts and to test and quantify the heterogeneity in random measurement error using Boot-Err.

For illustration, we analyse the Conscientiousness subscale of the HEXACO-personality inventory as administered in the multi-lab replication project reported by Verschuere et al. [-@RRRMazar]. In this project, Experiment 1 from Mazar et al. (ref.) was replicated across 19 independent laboratories. Each lab administered the same materials, including the HEXACO-inventory, to different participant samples scattered across the globe. This data-set therefore provides an ideal setting for Boot-Err: the same scale was administered across multiple independent samples under comparable but not identical conditions.

Importantly, our interest lies not in the experimental manipulation itself, but in the measurement properties of the HEXACO-Conscientiousness scale across laboratories. The HEXACO personality inventory was administered as a filler instrument between experimental tasks - accordingly, scores of Conscientiousness should not be affected by the experimental manipulation. Each lab constitutes a separate administration context, allowing us to examine whether random measurement error varies across contexts.

#### Tutorial structure

This tutorial provides a step-by-step guide on how to investigate differences in measurement quality across administration contexts using Boot-Err. All analyses are conducted in R. The tutorial includes executable chunks that can be run directly on the reader's own machine by copying the code into R. A quarto-file for this tutorial, containing both text and the executable code chunks can alternatively be found at \[.github / OSF-link\]. No additional proprietary software or supplementary materials are required. The data used within this tutorial is publicly available, we make use of an R-package to download and read the data-set. The entire procedure can be reproduced using the code presented in this section.

The tutorial is structured in 7 different steps: (0) data preparation, (1) estimating score reliability, (2) estimating error score variance, (3) using bootstrapping to generate standard errors, (4) meta-analysis of error score variance, (5) backtransforming estimates, (bonus) relating error score variance to total score variance.

At the end of this tutorial, readers will be able to

-   estimate error score variance for each administration context,

-   approximate the sampling variability of transformed error score variance using bootstrapping,

-   test whether error score variance differs across administration contexts, and

-   evaluate the magnitude of these differences in substantive terms

### Step 0 - Preparing Data

Since each data-set is different, depending on how data was collected and possibly aggregated, Step 0 is not formalized in the Boot-Err approach. Instead, the data needs to be cleaned and appropriately manipulated beforehand.

The following code chunk manages the packages required to follow this tutorial. By executing the code below, the user checks whether the required packages (magrittr, dplyr, osfr, here, zip, data.table, metafor, boot) are already installed. If the packages are not yet installed, they will be downloaded and installed from the default CRAN-repository. All required packages are subsequently loaded.

```{r, output = F, warning = F}
# relevant libraries required for this script
packages <- c("magrittr", "dplyr", "osfr", "here", "zip", "data.table", "metafor", "boot")

# check, whether library already installed or not - install and load as needed:
apply(as.matrix(packages), MARGIN = 1, FUN = function(x) {
  
  # check if package is installed on system
  pkg_avail <- nzchar(system.file(package = x))   
  
   # load the package, if already installed
  if(pkg_avail){
    require(x, character.only = TRUE)            
    
  # install the package, if missing
  }else{
    install.packages(x)      
    
    # load after installation
    require(x, character.only = TRUE)             
  }
})
```

After loading the required packages, we can turn towards downloading the data-set from the OSF-repository. While the data cleaning and manipulation procedure is bound to be different for every research project, we briefly discuss how we come to a final data-set for this tutorial. Firstly, we download the zipped replication data-set from the openly available OSF-repository (osf.io/fwnc2), using the osfr R package. After unzipping, we can read the data-file in R.

```{r, warning = F, eval = F}
# create relevant directories, so data can be downloaded and stpred
dir.create(here("Data"), showWarnings = FALSE)
dir.create(here("Data/Downloaded Data"), showWarnings = FALSE)


# use osfr-package to download RRR10 data on the Mazar et al. / Srull et al. replications
osfr::osf_retrieve_file("https://osf.io/fwnc2") %>% 
  osfr::osf_download(path = here("Data/Downloaded Data/"), conflicts = "overwrite")
# zip-file is stored at Downloaded Data

# again, unzip the file and move it to the corresponding directory
unzip(here("Data/Downloaded Data/Meta-Analysis_2018-07-09.zip"),
      files = "Meta-Analysis/Results/raw_data_corrected_MAA.csv",
      exdir = here("Data/Original Data/RRR10"),
      junkpaths = T)
```

For the purpose of this tutorial, we want to restrict our analyses to only those participants, which were included in the replicators' primary analyses. Therein, we end up with 4,674 participants from 19 laboratories. We achieve this in substep *a)* by identifying the relevant laboratories from the published manuscript and filtering out all laboratories that do not belong to the primary replication.

```{r, warning = F}
# HEXACO Conscientiousness

# a) Load data and select laboratories part of primary replication

# Thankfully, for the RRR10 project, the data can be taken from a file directly, without the need of combination:
pc_df <- as.data.frame(fread(here("Data/Original Data/RRR10/raw_data_corrected_MAA.csv")))

# data <- data[which(data$age >= 18 & data$age <= 25),]
# retain observations only for participants eligible for analysis
pc_df <- pc_df[which(pc_df$inclusion == "inclusion both RRR" | pc_df$inclusion == "inclusion Mazar only"),] %>%
  mutate(source = lab.name)

# identify relevant labs for analsysis
labs_in_paper <- c("Laine", "klein Selle & Rozmann", "Aczel", "Ferreira-Santos", "Meijer", "Loschelder", "Wick", "Suchotzki", 
                   "Sutan", "Vanpaemel", "Verschuere", "Wiggins", "Gonzalez-Iraizoz", "Koppel", "Birt", "McCarthy", "Evans", 
                   "Holzmeister", "Ozdogru")
labs_in_data <- unique(pc_df$source)
labs_in_data[8] <- "Gonzalez-Iraizoz"
labs_in_data[16] <- "Ozdogru"

# remove labs from data, which we do not need for analysis
labs_excl <- labs_in_data[!labs_in_data %in% labs_in_paper]
pc_df <- pc_df[which(!pc_df$source %in% labs_excl),]
```

The data-set contains a large amount of information, that we do not need, in the form of columns not required for this analysis. Therefore, in substep *b)* we retain only those columns from the data-set, that contain information about the participants' responses to the HEXACO-items and in which laboratory the participant was assessed. Since some of the HEXACO-items are reverse-coded, we need to re-code half of the items, so the scale is identical across all items.

```{r, warning = F}
# b) Select relevant variables and re-code items that are coded in the opposite direction

# retain only those columns, which are related to the HEXACO-personality inventory
# we can do so by only retaining columns, whose names start with "hex"
# As we need to differentiate between scores from laboratories, we keep column "source"
pc_df <- pc_df[,c(which(names(pc_df) %in% c("source")),
                  grep("^hex", names(pc_df)))]

# recoding the hexaco items, that need recoding

for(i in grep("^hex", names(pc_df))){
  pc_df[,i] <- as.integer(pc_df[,i])
}

# these are the numbers of the items, that need recoding
items_hex_recode <- c(30, 12, 60, 42, 24, 28, 53, 35, 41, 59, 28, 52, 10, 46, 9, 15, 57, 21, 26, 32, 14, 20, 44, 56, 1, 31, 49, 19, 55, 48)

# combining "hex" and number gives the column names
names_items_hex_recode <- paste0("hex", items_hex_recode) 

# adding _R for names of items, that are recoded
names_items_hex_recode_R <- paste0(names_items_hex_recode, "_R") 

# replace the original (un-recoded) scores with the recoded ones
pc_df[,names_items_hex_recode_R] <- 6 - pc_df[,names_items_hex_recode] 
```

Lastly, in substep *c)*, we remove all columns that contain responses on the other HEXACO-dimensions, which are not Conscientiousness. Therein, we have constructed a data-set, containing the responses on Conscientiousness from all participants that are part of the primary replication sample in Verschuere et al. (ref.).

```{r, warning = FALSE}
# c) select items belonging to Conscientiousness subscale

# Conscientiousness
# items in Conscientiousness subscale
items_hex_CO <- c(2, 26, 8, 32, 14, 38, 50, 20, 44, 56) 

# did item need recoding?
names_items_hex_CO <- ifelse(items_hex_CO %in% items_hex_recode, paste0("hex", items_hex_CO, "_R"), paste0("hex",items_hex_CO)) 

# select all items from Conscientiousness subscale, correctly coded
pc_hex_items_CO <- which(names(pc_df) %in% names_items_hex_CO) 

# retain only columns concerning Conscientiousness and source
pc_df <- pc_df[,c(names_items_hex_CO, "source")]

# identify individual laboratories
labs <- unique(pc_df$source)
```

However, we find that some participants' responses on the Conscientiousness-items are missing in this data-set. While this was of no concern to the replication-authors, as the HEXACO-inventory served as a filler instrument, the same is not true for this tutorial. We will use Cronbach's Alpha to estimate score reliability. Cronbach's Alpha can not deal with missing responses. To keep it simple for this tutorial, we make use of list-wise deletion, removing any missing responses from the data-set. This means that the data for each participant where any response on one or several items is missing is deleted from the data-set. Alternative approaches, that attempt to impute the missing responses, while retaining the general covariance-structure of the data-set, are available. Great examples include Multiple Imputation (ref.) or Multivariate Imputation by Chained Equations (ref.). Single value imputation methods like Mean or Mode Imputation restrict the data-set's covariance structure, leading to estimation issues and biases in subsequent analyses.

In the following code chunk, we compute how many scores are missing for each individual item. Subsequently, we retain only scores from participants, where no single item score for Conscientiousness is missing.

```{r, warning = FALSE}
# d) dealing with missing responses

# compute number of missing values per item
colSums(is.na(pc_df))

# retain only complete observation sets
pc_df <- na.omit(pc_df)
```

Finishing step 0, we have constructed a data-set, containing only the complete response sets of participants belonging to the replication authors' primary analysis sample. Subsequently, we can begin to follow the Boot-Err procedure.

### Step 1 - estimating score reliability

The first step in the Boot-Err procedure aims to generate estimates of score reliability for each sample or administration site respectively. In the following code, we use *sapply()* to compute Cronbach's Alpha separately for each laboratory in the data-set. Cronbach's Alpha is generally defined as outlined in the following @eq-alpha_formal (ref.).

$$\alpha=\frac{k^2\ \bar{\sigma}_{ij}}{\sigma^2_x}$$ {#eq-alpha_formal}

Here, $k$ corresponds to the number of items in the scale, $\bar{\sigma}_{ij}$ constitutes the average covariance between items and $\sigma^2_x$ constitutes the total score variance in the total test results.

However, computing the average item covariance can be computationally intensive, and a more efficient computation technique is available in @eq-alpha_fast (ref.).

$$\alpha = \frac{k}{k-1}\left(1-\frac{\sum_{i=1}^k\sigma^2_{i}}{\sum_{i=1}^k\sum_{j=1}^k\ \sigma_{ij}}\right) $$ {#eq-alpha_fast}

Here, $\sigma^2_{i}$ describes the variance of scores on item $i$, while $\sigma_{ij}$ describes the covariance between items $i$ and $j$. The equation in @eq-alpha_fast generates the same estimate of Cronbach's Alpha as in @eq-alpha_formal, but much more efficiently. By using *sapply* to quickly compute across all laboratories, we can efficiently generate a vector containing estimates of Cronbach's Alpha for each administration site respectively.

```{r, warning = F}
# compute Cronbach's Alpha across all administration contexts

# we use sapply as it tends to be faster than a for-loop
alpha_vec <- sapply(labs, FUN = function(lab){
  
  # select data for each administration context separately
  dat <- pc_df %>% 
    filter(source == lab) %>% 
    select(-"source")
  
  # compute covariance matrix
  C <- cov(dat)
  
  # extract the number of items
  j <- dim(C)[1]
  
  # compute Cronbach's Alpha efficiently
  alpha <- (1 - sum(diag(C))/sum(C)) * (j/(j - 1))
  
  return(alpha)
})

# return the estimate vector (rounded to 2 decimals)
round(alpha_vec, 2)
```

### Step 2 - estimating error score variance

After generating estimates of score reliability using Cronbach's Alpha, we can turn towards Step 2 of the Boot-Err procedure - estimating error score variance. In @eq-errorvar we established that error score variance can be constructed from a reliability coefficient and the total score variance. This means, provided we generate an estimate of total score variance, we can use @eq-errorvar to estimate error score variance, as we already have estimate of score reliability in the form of Cronbach's Alpha readily available.

In the following code chunk we, again, use an *sapply*-function to efficiently repeat the same computation for each laboratory. We estimate the total score variance $\sigma^2_X$ by computing Conscientiousness scores as the average item score for each participant (generating the test score) and computing the variance of these test scores. Subsequently, we can easily use @eq-errorvar to generate a vector containing error score variance $\sigma^2_E$-estimates for each laboratory respectively.

```{r}
# estimate error score variance for each administration context

# we use sapply as it tends to be faster than a for-loop
var_X_vec <- sapply(labs, FUN = function(lab){
  
  # select data for each administration context separately
  dat <- pc_df %>% 
    filter(source == lab) %>% 
    select(-"source")
  
  # compute the total score variance from individual test scores
  var_X <- dat %>% 
    rowMeans(.) %>% 
    var(.)
  
  return(var_X)
  
})

# compute error score variance from the estimates of reliability and total score variance
var_E_vec <-  (1-alpha_vec) * var_X_vec

# return the estimated vector (rounded to 3 decimals)
round(var_E_vec, 3)

```

### Step 3 - construct Standard Errors using Bootstrapping

As described in the introductory *Boot-Err*-section, we use a random-effects meta-analysis to estimate and test for heterogeneity in error score variances. Since estimation of total score variance and its components is not variance-stable, we need to use a variance-stabilizing transformation. Bartlett and Kendall [-@bartlett1946] show that using the natural log, as in @eq-stable_var, is a suitable variance-stabilizing transformation for total score variance estimates.

$$ T[\hat{\sigma}^2_x] = ln \left(\hat{\sigma}^2_x \right) $$ {#eq-stable_var}

While little evidence on variance-stabilizing transformations for score variance components is available, we have used the natural log for both the total and the error score variance component in our proposal for the Boot-Err [@beinhauer2025]. In the simulation scheme, we find a small but negligible bias in the estimates. We concluded that the transformation described in @eq-stable_var is suitable for the variance component of error score variance $\sigma^2_e$. Therefore, we are confident in our claim that @eq-stable_varE describes a suitable variance-stabilizing transformation.

$$ T[\hat{\sigma}^2_e] = ln \left(\hat{\sigma}^2_e \right) $$ {#eq-stable_varE}

However, in order to run a random-effects meta-analysis, we not only need a variance-stable estimate, but also a Standard Error, quantifying our uncertainty in that estimate. While we are aware of no analytic approximation for the Standard Error of log-transformed error score variance, we can use Bootstrapping techniques to generate robust Standard Error estimates for $T[\hat{\sigma}^2_e]$.

The package *boot* provides easy-to-use and efficient functions (including parallelization if desired) for Bootstrapping techniques. During the conception of this tutorial, we made use of version 1.3-31. In order to use *boot*, we need to provide the package with a user-written function, which needs to do two things: i) allow the *boot*-package to sample from our data-set (we do so by `dat_boot <- data[indices,]`), and ii) compute the parameter on the re-sampled data-set that we want to boot-strap. In the following code-chunk, we show a function that does just that.

```{r}
# define the user-written function to compute transformed error score variance

ln_var_E_boot_func <- function(data, indices){
  
  # essential for any boot-function: allow boot to sample from dataset rows
  dat_boot <- data[indices,]
  
  # we have to compute Cronbach's Alpha again for each bootstrapped sample
  
  # compute covariance matrix
  C <- cov(dat_boot)
  
  # extract number of items
  j <- dim(C)[1]
  
  # compute Cronbach's Alpha
  alpha <- (1 - sum(diag(C))/sum(C)) * (j/(j - 1))
  
  # we also have to compute error score variance again for each bootstrapped sample
  
  # compute total score variance
  var_X <- dat_boot %>% 
    rowMeans(.) %>% 
    var(.)
  
  # compute error score variance
  var_E <- var_X * (1 - alpha)
  
  # log-transform as variance stabilising transformation
  ln_var_E <- log(var_E)
  
  return(ln_var_E)
  
}
```

Now that we have a function available for *boot* to re-sample our data and estimate $T[\hat{\sigma}^2_e]$ in the bootstrapped samples, we need to supply our user-defined function to *boot* and compute the Standard Error from the bootstrapped estimates. Essentially, what we do here is estimate the Standard Error of log-transformed error score variance, by computing the standard deviation of all bootstrapped estimates for a single laboratory, see @eq-bootSE.

$$ SE\left[T[\hat{\sigma}^2_e]\right] = \sqrt{\frac{1}{n}\sum_{i=1}^n(T[\hat{\sigma}^2_e]_i - \bar{T}[\hat{\sigma}^2_e])^2}$$ {#eq-bootSE}

Here, $n$ describes the number of bootstrap-samples drawn, whereas $\bar{T}[\hat{\sigma}^2_e]$ describes the mean of the log-transformed error score variance estimates of the $n$ bootstrap samples. In the following code-chunk, we use the *boot* package to estimate $SE\left[T[\hat{\sigma}^2_e]\right]$ from 3000 bootstrap samples for each laboratory respectively. Again, we make use of an *sapply*-function to efficiently compute across all laboratories.

```{r, warning = FALSE}
# approximate the standard error for error score variance

ln_var_E_SE_vec <- sapply(labs, FUN = function(lab){
  
  # select data for each administration context separately
  dat <- pc_df %>% 
    filter(source == lab) %>% 
    select(-"source")
  
  # use boot and our user-defined function to efficiently generate transformed error score variance in 3000 bootstrapped samples
  boot_estimates <- boot::boot(dat, 
                               statistic = ln_var_E_boot_func,
                               R = 3000)
  
  # compute the standard error
  SE <- sd(boot_estimates$t)
  
  return(SE)
  
})
```

Lastly, we still need to formally transform and store the individual estimates of $T[\hat{\sigma}^2_e]$ for each laboratory (so far we only did this for the bootstrap-samples). However, at this point, computing this transformation is trivial. To make things easier for us in the following steps, we combine the estimates of log-transformed error score variance and its Standard Errors here in a single data-frame. We add the laboratory names to the data-frame, so estimates can be linked to their respective sources.

```{r, warning = FALSE}
# combine standard errors and estimates in a data-frame

# transform individual estimates of error score variance
ln_var_E_vec <- log(var_E_vec)

# combine estimates, standard error and lab-name in data-frame
ln_var_E_df <- data.frame(ln_var_E = ln_var_E_vec, 
                          ln_var_E_SE = ln_var_E_SE_vec, 
                          laboratory = labs, 
                          row.names = NULL)

# return the estimates and round the numeric variables (ln-estimate and SE)
ln_var_E_df %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

Concluding step 3, we have gathered all estimates we need, to perform a random-effects meta-analysis on (log-transformed) error score variances.

### Step 4 - meta-analysis of error score variance

For this step, we estimate the heterogeneity of log-transformed error score variance and test whether it can be statistically distinguished from zero. We use the R-package *metafor*, as it provides an efficient and easy-to-use implementation of random-effects meta-analyses for any parameter [@metafor]. Using the *rma*-function with the argument *method = "REML"* ensures that we use the REML-estimator of heterogeneity. Hönekopp and Linden [-@hoenekopp2022] have shown that the REML-estimator tends to outperform other estimators across a wide range of conditions.

The following chunk will run a random-effects meta-analysis in *metafor*, generating an estimate of the meta-analytic mean (in the output `estimate`), estimates of absolute heterogeneity in standard deviation and variance (in the output `tau` and `tau^2`) and estimates of relative heterogeneity in the form of $I^2$ and $H^2$ (in the output `I^2` and `H^2`). Additionally, *metafor* provides the results for two hypothesis tests. Firstly, the test for heterogeneity (`Test for Heterogeneity` in the output), assessing whether the estimate of heteorgeneity $\tau^2$ can be statistically distinguished from $0$. Secondly, the test for the meta-analytic effect size, assessing whether the meta-analytic aggregate can be statistically distinguished from $0$ (in the Output `Model results`).

```{r}
# perform the random-effects meta-analysis on log-transformed error score varice
ln_var_E_rma <- metafor::rma(data = ln_var_E_df,
                             yi = ln_var_E,
                             sei = ln_var_E_SE,
                             method = "REML")

ln_var_E_rma

```

The significance tests indicate two things for the log-transformed error score variance of Conscientiousness across replications in Verschuere et al. [-@RRRMazar]: Firstly, we find that the meta-analytic estimate of `r paste0("$\\mu_{T[\\hat{\\sigma}^2_e]} = ", round(ln_var_E_rma$b[1], 3), "$")` is significantly different from zero (`r paste0("$z = ", round(ln_var_E_rma$zval, 3), "$")`, `r paste0("$p ", ifelse(ln_var_E_rma$pval < .001, yes = paste0('< .001'), no = paste0('= ', round(ln_var_E_rma$pval, 3))), "$")`), using a significance level of 5%. However, this information is not particularly meaningful, the estimate is on the log-transformed scale and interpretation is not straightforward. Secondly and more interestingly, we find that the test for heterogeneity indicates that there is statistically significant heterogeneity in log-transformed error score variances (`r paste0("$Q(", ln_var_E_rma$k - 1, ") = ", round(ln_var_E_rma$QE, 3), "$")`, `r paste0("$p ", ifelse(ln_var_E_rma$QEp < .001, yes = paste0('< .001'), no = paste0('= ', round(ln_var_E_rma$QEp, 3))), "$")`). Even though the tested estimates are still on the log-scale, we can nonetheless infer from this result, that error score variance $\sigma^2_E$ is not identical across all administrations of the Conscientiousness-scale.

However, the estimate of `r paste0("$\\tau^2_{T[\\hat{\\sigma}^2_e]} = ", round(ln_var_E_rma$tau2, 3), "$")` can not be easily interpreted due to the log-transformation. Instead, the estimates of relative heterogeneity may be interpreted despite the transformation. The estimate of `r paste0("$I^2 = ", round(ln_var_E_rma$I2, 2), "$")` implies that `r round(ln_var_E_rma$I2, 2)`% of the variance observed in the estimate of error score variance are due to actual heterogeneity in the population, not due to sampling error. Similarly, the estimate of `r paste0("$H^2 = ", round(ln_var_E_rma$H2, 2), "$")` implies that the heterogeneity is `r round(ln_var_E_rma$H2, 2)` times as large as the variance induced by sampling error. According to standard conventions (ref.), both estimate imply a strong or large extent of heterogeneity. However, both estimates of relative heterogeneity are often criticized, as they depend strongly on sample size and are easily manipulated (refs.).

Instead, we should turn towards an informative interpretation of the absolute heterogeneity in terms of $\tau^2$ or $\tau$. To do so, we need to back-transform our newly generated estimates to the original scale. We do so in Step 5.

### Step 5 - Back-transforming estimates

The transformation of score variance (components) using the natural log not only serves to generate variance-stable estimates. Additionally, it comes with the added benefit, that we can assume $T[\sigma^2_e]$ to follow a normal distribution. Several authors have proposed that the total score variance, if not constant, can be described by a log-normal distribution (refs.). This means that the log-transformed estimates accordingly follow a normal distribution. This is very helpful for the application of the Boot-Err technique, as one central assumption of a random-effects meta-analysis is that the parameter follows a normal distribution. Additionally, it helps us back-transform the estimates of meta-analytic mean $\mu_{T[\sigma^2_e]}$ and its heterogeneity $\tau^2_{T[\sigma^2_e]}$, as equations are readily available (refs.).

$$ \hat{\mu}_{\sigma^2_e} = \exp\left(\hat{\mu}_{T[\sigma^2_e]} + \frac{1}{2} \hat{\tau}^2_{T[\sigma^2_e]}\right) $$ {#eq-backtransf_mu}

@eq-backtransf_mu demonstrates how the mean error score variance $\hat{\mu}_{\sigma^2_e}$ can be estimated from the meta-analytic estimates of $T[\sigma^2_e]$. Here, $\hat{\mu}_{T[\sigma^2_e]}$ represents the estimate of meta-analytic mean of log-transformed error score variance and $\hat{\tau}^2_{T[\sigma^2_e]}$ describes the estimate of its heterogeneity. Similarly, @eq-backtransf_var can be used to derive the heterogeneity of error score variance $\hat{\tau}^2_{\sigma^2_e}$ on its original scale, from the same meta-analytic estimates of \$T\[\sigma\^2_e\]\$\$.

$$ \hat{\tau}^2_{\sigma^2_e} = \exp\left(\hat{\tau}^2_{T[\sigma^2_e]} - 1\right) \cdot \exp\left(2 \hat{\mu}_{T[\sigma^2_e]} + \hat{\tau}^2_{T[\sigma^2_e]}\right) $$ {#eq-backtransf_var}

In the following chunk, we define two functions which compute [Equations -@eq-backtransf_mu] and [-@eq-backtransf_var] from the *metafor*-output. The functions only take the *metafor*-object generated from the *rma*-function as input and return the estimates $\hat{\mu}_{\sigma^2_e}$ and $\hat{\tau}^2_{\sigma^2_e}$ respectively.

```{r}
# two functions to back-transform heterogeneity and meta-analytic estimate

# back-transform meta-analytic estimate (the weighted mean)
backtransform_var_mu <- function(rma_obj){
  exp(rma_obj$b[1] + (.5*rma_obj$tau2))
}

# back-transforming meta-analytic heterogeneity
backtransform_var_tau2 <- function(rma_obj){
  (exp(rma_obj$tau2) - 1) * exp((2 * rma_obj$b[1]) + rma_obj$tau2)
}


round(backtransform_var_mu(ln_var_E_rma), 3)

round(backtransform_var_tau2(ln_var_E_rma) %>% 
  sqrt(.), 3)
```

Running the two functions in the chunk above returns two estimates: The meta-analytic mean error score variance of the HEXACO-Conscientiousness scale across 19 laboratories is $\hat{\mu}_{\sigma^2_e} = `{r} round(backtransform_var_mu(ln_var_E_rma), 3)`$; The estimate for its heterogeneity is $\hat{\tau}_{\sigma^2_e} =`{r} round((backtransform_var_tau2(ln_var_E_rma) %>%   sqrt(.)), 3)`$. To understand what this implies, we can assess the relative size of heterogeneity, also known as the coefficient of variation, which is computed as $CV = \frac{\tau}{\mu}$. Here we find a CV of $CV_{\sigma^2_e} = `{r} round((backtransform_var_tau2(ln_var_E_rma) %>% sqrt(.))/backtransform_var_mu(ln_var_E_rma), 3)`$. This means that the extent of the differences in error score variances is approximately `{r} round((backtransform_var_tau2(ln_var_E_rma) %>% sqrt(.))/backtransform_var_mu(ln_var_E_rma), 3) * 100`% of the mean error score variance found across these laboratories. Based on the random-effects model, true error score variance across administrations can be expected to lie approximately between X and Y. These values can be obtained by adding and subtracting the estimated heterogeneity $\hat{\tau}_{\sigma^2_e}$ from the meta-analytic mean $\hat{\mu}_{\sigma^2_e}$ (on the transformed scale) and subsequently back-transforming to the original variance metric. For Conscientiousness, we find an interval where we expect the error score variance to vary at least between `r round(exp(ln_var_E_rma$b[1] - sqrt(ln_var_E_rma$tau2)), 3)` and `r round(exp(ln_var_E_rma$b[1] + sqrt(ln_var_E_rma$tau2)), 3)`

None of the other approaches, such as reliability generalization meta-analysis or measurement invariance testing allow to generate these estimates we have derived using the Boot-Err technique. For researchers who have experience working with the respective scale, these estimates can be highly valuable to understand how much variance in scores we can expect to be introduced by measurement imprecision across multiple administrations of the same scale. While we, as authors, lack the expertise on what to expect from the conscientiousness scale, we can construct additional parameters to gain a deeper understanding. In the following Bonus-Step, we delve deeper into the insights generated by the Boot-Err technique.

### Bonus-Step - Relating error score variance to total score variance

Score reliability $\rho_{xx'}$ is defined as the relative extent to which total score variance observed is attributable to actual differences in the true scores, not to measurement imprecision, see @eq-reliability. However, that also means that $1 - \rho_{xx'}$ describes the relative extent to which total score variance is attributable to random measurement error.

Similarly, on the meta-analytic scale, we can look at the heterogeneity in total score variance $\sigma^2_x$, and attempt to understand how much of its heterogeneity can be attributed to differences in error score variance. Intuitively, many researchers tend to ascribe heterogeneity in total score variance $\sigma^2_x$ to differences in measurement precision. This means that often, a larger score variance is interpreted as if it was caused by larger measurement error such as $\sigma^2_e$. However, substantial parts of said heterogeneity might also be better explained by differences in the true score variance $\sigma^2_t$. Fortunately, if we follow Steps 3 through 5 for the total score variance $\sigma^2_x$ instead, we can assess in how far its heterogeneity can be attributed to differences in true score variance $\sigma^2_t$ or measurement error $\sigma^2_e$.

In the following chunk, we briefly run through these steps to estimate heterogeneity in total score variance $\sigma^2_x$.

```{r}
# perform steps 3 through 5 for total score variance

# define the function for boot to bootstrap transformed total score variance
ln_var_X_boot_func <- function(data, indices){
  
  # allow boot to resample from data-set
  dat_boot <- data[indices,]
  
  # compute total score variance
  var_X <- dat_boot %>% 
    rowMeans(.) %>% 
    var(.)
  
  # transform total score variance
  ln_var_X <- log(var_X)
  
  return(ln_var_X)
}


# use Bootstrapping to compute standard error of log-transformed total score variance
ln_var_X_SE_vec <- sapply(labs, FUN = function(lab){
  
  # select data only for specific administration context
  dat <- pc_df %>% 
    filter(source == lab) %>% 
    select(-"source")
  
  # use boot to bootstrap log-transformed total score variance
  boot_estimates <- boot::boot(dat, 
                               statistic = ln_var_X_boot_func,
                               R = 3000)
  
  # estimate standard error
  SE <- sd(boot_estimates$t)
  
  return(SE)
  
})

# transforme total score variance estimates
ln_var_X_vec <- log(var_X_vec)

# combine estimates of total score variance and its standard error in single df
ln_var_X_df <- data.frame(ln_var_X = ln_var_X_vec, 
                          ln_var_X_SE = ln_var_X_SE_vec, 
                          laboratory = labs, 
                          row.names = NULL)

# run random-effects meta-analysis
ln_var_X_rma <- metafor::rma(data = ln_var_X_df,
                             yi = ln_var_X,
                             sei = ln_var_X_SE,
                             method = "REML")

# round back-transformed meta-analytic estimate to 3rd decimal
round(backtransform_var_mu(ln_var_X_rma), 3)

# round back-transformed heterogeneity estimate to 3rd decimal
round(backtransform_var_tau2(ln_var_X_rma) %>% 
  sqrt(.), 3)

```

We find a meta-analytic mean of total score variance $\hat{\mu}_{\sigma^2_x} = `{r} round(backtransform_var_mu(ln_var_X_rma), 3)`$ and a heterogeneity of $\hat{\tau}_{\sigma^2_x} = `{r} round((backtransform_var_tau2(ln_var_X_rma) %>% sqrt(.)), 3)`$. Under CTT, total score variance is defined as the sum of true and error score variance $\sigma^2_x = \sigma^2_e + \sigma^2_t$, which are assumed to be independent. This means that we can assess how much heterogeneity in total score variance can actually be explained by differences in measurement error:

$$ R = \frac{\hat{\tau}^2_{\sigma^2_e}}{\hat{\tau}^2_{\sigma^2_x}} $$ {#eq-ratio}

In @eq-ratio, we define the ratio-variable $R$ as the ratio of $\hat{\tau}_{\sigma^2_e}$ to $\hat{\tau}_{\sigma^2_X}$. For Conscientiousness, we find a ratio of $R  = `{r} round((backtransform_var_tau2(ln_var_E_rma)) / (backtransform_var_tau2(ln_var_X_rma)), 3)`$. This means that about `{r} round((backtransform_var_tau2(ln_var_E_rma)) / (backtransform_var_tau2(ln_var_X_rma)), 3) * 100`% of heterogeneity in $\sigma^2_x$ could be explained by differences in random measurement error. This also means that `{r} round((1 - (backtransform_var_tau2(ln_var_E_rma)) / (backtransform_var_tau2(ln_var_X_rma))), 3) * 100`% of this heterogeneity is explained by differences in true score variance $\sigma^2_t$.

By following the steps from 1 through 5 we have assessed differences in measurement quality of Conscientiousness across different administration contexts. In doing so, including the Bonus-Step, we have generated a number of estimates and parameters that help us understand in how far these differences are substantive. In the following section, we turn towards a substantive interpretation of what these results imply for the use of the HEXACO-Conscientiousness inventory.

### Conclusion

Boot-err essentially consists of 5 steps, where we estimate score reliability (step 1), estimate error score variance (step 2), use appropriate variance-stabilizing transformations paired with Bootstrapping to estimate standard errors (step 3), perform meta-analyses of error score variance (step 4) and back-transform estimates back to the original scale (step 5). In the previous section, we provided custom functions in R for researchers to perform a Boot-Err analysis on data obtained from a single measuring instrument administered across different contexts. Combining the individual steps allows for an assessment of whether there is heterogeneity in random measurement error across administrations of the same scale.

Making use of the data supplied by Verschuere et al. [-@RRRMazar], we used Boot-Err to understand in how far there are differences in absolute measurement precision across 19 laboratories where the HEXACO-Conscientiousness personality scale was administered. We not only find statistically significant differences in error score variance, but were also able to estimate the absolute and relative extent of these differences. Therein, we are certain that the random measurement error in the Conscientiousness scores is not constant across samples. We find that the heterogeneity of `{r} round(backtransform_var_tau2(ln_var_E_rma), 3)` is about `{r} round((backtransform_var_tau2(ln_var_E_rma) %>% sqrt(.))/backtransform_var_mu(ln_var_E_rma), 3) * 100`% of the mean error score variance. Across administrations of the same scale, we can expect error score variance to vary at least between `r round(exp(ln_var_E_rma$b[1] - sqrt(ln_var_E_rma$tau2)), 3)` and `r round(exp(ln_var_E_rma$b[1] + sqrt(ln_var_E_rma$tau2)), 3)`.

In order to better understand what these differences mean, we additionally uncovered in how far the heterogeneity in total score variance can be attributed to these differences in measurement error. Here, we found that the overwhelming majority of heterogeneity, `{r} round((1 - (backtransform_var_tau2(ln_var_E_rma)) / (backtransform_var_tau2(ln_var_X_rma))), 3) * 100`%, can not be explained by differences in measurement error. For the Conscientiousness-scale, this implies that, while there clearly are substantial differences in measurement error across groups, the distribution of true scores varies much more strongly across groups. This is important, as this implies that the differences in total score variance can not be attributed to differences in measurement error. Instead, Conscientiousness is dispersed highly different across those groups. In Beinhauer et al. [-@beinhauer2025liable] we discuss potential causes and implications of this phenomenon.

## Discussion

Measurement precision and how strongly it diverges across administrations of the same instrument is a central question in social and behavioural sciences. Differences in random measurement error are not only interesting to researchers assessing the quality of a measuring instrument. These differences also have clear implications concerning hypothesis testing and standardization of effect sizes. We have summarised some of the most popular techniques assessing differences in random measurement error - reliability generalization meta-analysis and measurement invariance testing. Additionally, we introduced Boot-Err, a technique we developed to step away from relative measures such as score reliability or tests without estimates of how large the differences are, like measurement invariance tests [@beinhauer2025].

In the tutorial-section, we have demonstrated how Boot-Err can be used to generate detailed estimates and tests regarding the extent of the differences we observe in random measurement error, and how strongly they affect the differences in score variance. It is important to note that this technique comes with a number of limitations and constraints to generalizability. Just like for any statistical analysis, a number of assumptions need to be met. Regarding the application of Boot-Err, total and error score variance should follow a log-normal distribution and be computed from samples that are generally independent of each other (the individual samples should not be described as if stemming from a nested structure). Additionally, a number of assumptions relate to the type of score reliability estimate used. Here, we made use of Cronbach's Alpha, which typically comes with assumptions of tau-equivalence and unidimensionality (ref.). While these assumptions are widely criticized and rarely hold up in real data (ref.), we need to remind ourselves that these assumptions also relate to typical scoring methods of these measuring instrument. The HEXACO-personality score for each dimension is computed by taking the average of the relevant item scores (ref. Hexaco). Therein, even stricter assumptions going beyond tau-equivalence and unidimensionality apply - making Cronbach's Alpha the more appropriate choice of score reliability estimate over other ones that may impose less strict assumptions.

Even though differences in absolute measurement precision are of central interest to measuring instrument and scale development, methods to assess differences in error score variance each come with their own caveats and shortcomings. Models regarding differences in score reliability fail to take differences introduced by true score variance into account, while models simply testing for differences in error score variance without actually estimating it fail to put these differences into perspective. Perhaps this is the reason why discussions surrounding the impact of random measurement error on standardized effect sizes and hypothesis tests has been overemphasized, implying that removing random measurement error was akin to harmonizing variability across populations and data-sets.

Boot-Err, mends both these shortcomings, as differences in error score variance can be estimated and tested, but also put into perspective on how much heterogeneity in total score variance they actually produce. As we demonstrated in the Bonus-Step section, differences in error score variance can not explain why total score variance differs across groups--instead, differences in true score variance must come at a much larger extent. This implies that differences in score variability can not sufficiently be mended by taking care of measurement error. Instead, future research should ensure that the true score variance is stable across groups and representative for the respective population.

## References

::: {#refs}
:::
